---
title: "The Odd-Toed Ungulates Example"
author: "Guillaume Guénard"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
bibliography: ../inst/bib/Odd_toad_ungulates.bib
vignette: >
  %\VignetteIndexEntry{The Odd-Toed Ungulates Example}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})
##
### Load packages here:
##
### Figure counter:
(
  function() {
    log <- list(
      labels = character(),
      captions = character()
    )
    list(
      register = function(label, caption) {
        log$labels <<- c(log$labels, label)
        log$captions <<- c(log$captions, caption)
        invisible(NULL)
      },
      getNumber = function(label) {
        which(log$labels == label)
      },
      getCaption = function(label) {
        a <- which(log$labels == label)
        cap <- log$captions[a]
        cat(sprintf("Figure %d. %s\n\n---\n",a,cap))
        invisible(NULL)
      }
    )
  }
)() -> figCounter

(
  function() {
    log <- list(
      labels = character(),
      captions = character()
    )
    list(
      register = function(label, caption) {
        log$labels <<- c(log$labels, label)
        log$captions <<- c(log$captions, caption)
        invisible(NULL)
      },
      getNumber = function(label) {
        which(log$labels == label)
      },
      getCaption = function(label) {
        a <- which(log$labels == label)
        cap <- log$captions[a]
        cat(sprintf("Table %d. %s\n\n---\n",a,cap))
        invisible(NULL)
      }
    )
  }
)() -> tabCounter
```

Package version: `r packageVersion("RPEM")`


## Introduction

We prepared adapted version of the odd-toed ungulates (Mammalia: Perissodactyla)
---originally published by @PurvisRambaut1995 in conjunction with the
Comparative Analysis by Independent Contrasts (CAIC) software
[@PurvisRambautCAIC]---which we used for illustrating PEM calculations and
potential benefits. We began by obtaining the original dataset via the **caper**
package in R. It includes trait data for 13 species of odd-toed ungulates,
specifically mean female body mass, mean gestation length, mean neonatal body
mass, and a binary indicator of territorial behavior.

```{r echo=FALSE}
tabCounter$register(
  "GenBankData",
  paste(
    "GenBank accession numbers for the mitochondrial DNA sequences used to",
    "reconstruct the phylogeny of odd-toed ungulates, including two-letter",
    "species abbreviations.")
)
```

All continuous traits were provided in log-transformed form to improve normality
and facilitate comparisons across orders of magnitude. Although the original
dataset includes an associated phylogenetic tree, we did not use it. Instead, we
reconstructed a more up-to-date phylogeny for these species using mitochondrial
DNA sequences, leveraging the availability of complete or near-complete
mitogenomes in the National Center for Biotechnology Information (NCBI)
[GenBank](https://www.ncbi.nlm.nih.gov/genbank/) database
(Table `r tabCounter$getNumber("GenBankData") `).

---

```{r, echo=FALSE, results='asis'}
tabCounter$getCaption("GenBankData")
```

| Species                     | Accession   | Abbreviation |
| :-------------------------  | :--------:  | -----------: |
| _Dicerorhinus sumatrensis_  | NC_012684.1 | DS           |
| _Rhinoceros sondaicus_      | NC_012683.1 | RS           |
| _Rhinoceros unicornis_      | NC_001779.1 | RU           |
| _Diceros bicornis_          | NC_012682.1 | DB           |
| _Ceratotherium simum_       | NC_001808.1 | CS           |
| _Tapirus terrestris_        | NC_053962.1 | TT           |
| _Tapirus pinchaque_         | OQ420428.1  | TP           |
| _Tapirus indicus_           | NC_023838.1 | TI           |
| _Equus grevyi_              | NC_020432.2 | EG           |
| _Equus burchelli_           | NC_044858.1 | EB           |
| _Equus zebra_               | NC_020476.1 | EZ           |
| _Equus africanus_           | KX669267.1  | EA           |
| _Equus hemionus_            | NC_016061.1 | EH           |
| _Bos taurus<sup>1</sup>_    | NC_006853.1 |              |

1. The domestic ox, _Bos taurus_ was used as an outgroup to root the tree.

---

The aligned DNA sequences were concatenated, excluding the mitochondrial control
region and origin of replication. The resulting alignment was used to compute a
pairwise DNA distance matrix under the F81 model [@Felsenstein1981] using the
`dist.dna()` function from the **ape** package in R [@Paradis2004].

From this distance matrix, a minimum evolution (ME) tree was inferred using the
`fastme.ols()` function, which seeks to minimize the sum of squared differences
between the observed DNA distances and the patristic distances (i.e., path
lengths along the tree) between the tips [@Lefort2015].

Reticulations were added sequentially to the initial tree, guided by the same
minimum evolution criterion used during tree inference. At each step, all
possible pairs of edges were evaluated as candidate connections for
cross-linking. For each candidate pair, the optimal positions of the
reticulation nodes---i.e., the points of entry and exit along the respective edges---and the length of the reticulation edge were jointly estimated using the
L-BFGS-B optimization algorithm [@Byrd1995], which supports box constraints.
This ensured that the reticulation nodes remained within the bounds of their
host edges. The length of the reticulation edge was constrained to the interval
[0, 1] to maintain biological plausibility and numerical stability.

Among all evaluated candidates, the reticulation that yielded the smallest sum
of squared differences between the observed DNA distances and the (mean)
patristic distances in the graph was selected and added. The resulting
reticulated phylogenetic graph was then used as the starting point for the next
iteration. The process was stopped after ten reticulations had been added,
providing a plausible, albeit illustrative, example of a phylogenetic network.

The dataset---including the F81 DNA distance matrix, the initial phylogenetic
tree, a table detailing the estimated reticulation events, and trait data for
the 13 odd-toed ungulates---is distributed with the RPEM package as an external
data object. It can be loaded into the R workspace using the following command:

```{r message=FALSE}
## Load RPEM
library(RPEM)

## Loading the data:
system.file(
  "extdata/Perissodactyla.rds",
  package = "RPEM"
) %>%
  readRDS -> perisso
```


## Calculations

In this example, we aim to predict the trait values available in the trait data
using PEMs derived from $11$ distinct phylogenetic structures: the initial
phylogenetic tree and each of the ten phylogenetic graphs obtained by
sequentially adding reticulation events. This approach enables us to evaluate
how incorporating reticulate evolution affects trait prediction accuracy and
model performance.

```{r echo=FALSE}
figCounter$register(
  "phylogeny",
  paste(
    "The 11 phylogenetic graphs estimated in this study: the initial tree",
    "(Net0) and ten reticulation networks (Net1--Net10). Networks Net1 to",
    "Net10 include an increasing number of reticulation events, ranging from",
    "1 to 10. The sum of squared deviations between the F81 DNA distances and",
    "the mean patristic distances in each graph are shown in parentheses",
    "after the graph’s name."
  )
)
```

The minimum evolution tree is first converted into a graph object and stored in a list. Subsequently, cross-links are added sequentially—based on the inferred reticulation events–to generate a series of candidate phylogenetic graphs, as illustrated below (Figure  `r figCounter$getNumber("phylogeny")`):

```{r}
net <- list(Net0=as.graph(perisso$phy))

for(i in 2L:nrow(perisso$ret)) {
  crosslink(
    net[[i - 1L]],
    link = data.frame(
      length = perisso$ret[i,"length"],
      from = perisso$ret[i,"from"],
      loc_from = perisso$ret[i,"loc_from"],
      to = perisso$ret[i,"to"],
      loc_to = perisso$ret[i,"loc_to"],
      row.names = sprintf("Ret%d", i - 1L)
    )
  ) -> net[[sprintf("Net%d", i - 1L)]]
}
```

---

```{r, echo=FALSE, fig.width=10, fig.height=10}
par(mfrow=c(4,3), mar=c(1,1,1,1))

c(`Dicerorhinus sumatrensis` = "DS", `Rhinoceros sondaicus` = "RS",
  `Rhinoceros unicornis` ="RU", `Diceros bicornis` = "DB",
  `Ceratotherium simum` = "CS", `Tapirus terrestris`= "TT",
  `Tapirus pinchaque`= "TP", `Tapirus indicus` = "TI", `Equus grevyi` = "EG",
  `Equus burchelli`= "EB", `Equus zebra`= "EZ", `Equus africanus` = "EA",
  `Equus hemionus` = "EH") -> abbrev

plotgr0 <- function(x) {
  rownames(x)[x$species] <- abbrev[rownames(x)[x$species]]
  x
}

for(i in names(net))
  plotgr0(net[[i]]) %>%
    plot(xpd=TRUE, cex.lab=0.7, cex.min=3.5,
         edge.color=list("black", "black"),
         main=sprintf("%s (%f)",i, perisso$ret[i,"SSD"]))
```

```{r, echo=FALSE, results='asis'}
figCounter$getCaption("phylogeny")
```

---

Phylogenetic Eigenvector Map calculations are described in detail by
@Guenard2013 and are only briefly summarized here. A PEM is derived from the
influence matrix, a rectangular binary matrix of dimensions $V\timesE$, where
$V$ denotes the set of vertices and $E$ the set of edges in the phylogenetic
graph. Each entry in the matrix is $1$ if a given vertex is located after a
specific edge along the graph’s paths, and $0$ otherwise. This matrix can be
computed using the function `InflMat()`.

Each column of the influence matrix is assigned a weight $w_i$, calculated based
on the phylogenetic distance $d_i$ associated with the corresponding edge. The
weighting function proposed by @Guenard2013 is:

$$
w_i = \psi d_i^\frac{1-a}{2},
$$

where the selection parameter a controls the degree of deviation from neutral
trait evolution and takes values between $0$ (indicating completely neutral
evolution) and $1$ (indicating strongly non-neutral evolution). The parameter
$\psi$ represents a scaling factor for the overall phylogenetic variance and is
typically set to positive values centered around $1$.

After weighting the columns, the rows corresponding to extant species are
extracted according to their associated vertex. The resulting matrix is
column-centered to zero mean and decomposed via singular value decomposition
(SVD) to identify its principal axes of phylogenetic variation.

The left singular vectors obtained from the SVD serve as eigenvector maps that
capture the major patterns of phylogenetic variation. These vectors are commonly
used as predictors in linear models, such as multiple regression, to account for
non-independence due to shared evolutionary history. The right singular vectors,
when scaled by the inverse of the corresponding singular values, form a
projection matrix that enables the computation of phylogenetic scores at any
specified location within the phylogenetic graph, including along edges or at
reticulation nodes.

In RPEM, PEM calculation is performed using the function `PEM()`, which requires
at minimum a phylogenetic graph as input. A basic usage example is shown below:

```{r}
PEM(net$Net0)
```


The default call to `PEM()` computes the PEM under the assumption of neutral
trait evolution. Users can modify this assumption by specifying different levels
of selection through the `a` argument, which accepts logit-transformed values of
the selection parameter. Consequently, the default corresponds to `a = -Inf`,
representing the limit of purely neutral evolution.

To estimate the value of the selection parameter `a` from empirical data, RPEM
provides the `evolution.model()` function, which fits the parameters of the
weighting function using observed trait data within the PEM framework. This
function is applied as follows to each of the phylogenetic graphs in the series:

```{r}
## Lists storing the eigenvectors (PEM) and optimization information
pem <- list()
opt <- list()

for(i in names(net)) {
  pem[[i]] <- PEM(net[[i]], a = 0)
  opt[[i]] <- evolution.model(pem[[i]], y=perisso$trait$log.female.wt)
}

## The 'a' parameter estimated for each graph in Fig. 1
unlist(lapply(opt, function(x) invlogit(x$par$a)))
```

The eigenvectors derived from a PEM can be extracted using the functions
`as.matrix.PEM()` or `as.data.frame.PEM()`, which return the results in matrix
or data frame format, respectively. These eigenvectors can then be used as
predictors in other modeling frameworks to estimate trait values or account for
phylogenetic signal in statistical analyses.

Alternatively, the helper function `pemlm()` provides a streamlined approach for
modeling with PEM eigenvectors. In its current implementation, this function
facilitates the selection of a subset of eigenvectors based on minimizing the
Akaike information criterion corrected for small sample sizes [AICc,
@Burnham2002; @Guenard2013].

We illustrate the typical workflow of `pemlm()` by predicting the missing
(log-transformed) gestation length for the mountain tapir (_Tapirus pinchaque_;
TP) using the PEM eigenvectors derived from the initial phylogenetic tree
(Net0), together with female and neonatal weights as auxiliary predictor
variables. The workflow begins by determining the graph location of the mountain
tapir using the `locate()` function, preparing a suitable dataset with the
helper function `model.data()`, and then calculating and optimizing the PEM as
follows:

```{r}
## Calculate the graph location and residual graph
train <- locate(net$Net0, "Tapirus pinchaque")

## Prepare the data for evolution.model()
model.data(
    formula = log.gestation.length~log.female.wt+log.neonatal.wt,
    data = perisso$trait
  ) -> mdat

## Calculate the (raw) PEM with a finite value for argument `a`
pem.train <- PEM(train$x, a = 0)

## Optimize the PEM for the data set, excluding focal species TP
evolution.model(
    object = pem.train,
    y = mdat$y[names(mdat$y) != "Tapirus pinchaque"],
    x = mdat$x[rownames(mdat$x) != "Tapirus pinchaque",]
  ) -> opt
```

Using the optimized PEM thus obtained, the `pemlm()` function is called to fit
an initial model, which is then refined by the embedded `$forward()` method
through the sequential addition of PEM eigenvectors as follows:

```{r}
## Initiate the linear model
pemlm(
  formula = log.gestation.length~log.female.wt+log.neonatal.wt,
  data = perisso$trait %>% .[rownames(.) != "Tapirus pinchaque",],
  pem = pem.train
) -> lm1

## Add the eigenvectors to the model, returns the indices of the eigenvectors
## selected in the regression model
lm1$forward()
```

A summary of the resulting model can be obtained using the `summary()` method,
or an ANOVA table can be calculated using the `anova()` method as follows:

```{r}
## Calculate the summary
summary(lm1)

## Calculate the ANOVA table
anova(lm1)
```

Finally, the predicted gestation length is obtained using method `predict()` as
follows:

```{r}
## Predict the gestation length for the Mountain tapir
predict(
  object = lm1,
  newdata = perisso$trait["Tapirus pinchaque",],
  newloc = train$location
) -> prd
```

```{r, echo=FALSE, cache=TRUE}
runExample <- function(k) {
  
  train <- locate(net[[k]], "Tapirus pinchaque")
  
  model.data(
    formula = log.gestation.length~log.female.wt+log.neonatal.wt,
    data = perisso$trait
  ) -> mdat
  
  pem.train <- PEM(train$x, a = 0)
  
  evolution.model(
    object = pem.train,
    y = mdat$y[names(mdat$y) != "Tapirus pinchaque"],
    x = mdat$x[rownames(mdat$x) != "Tapirus pinchaque",]
  ) -> opt
  
  pemlm(
    formula = log.gestation.length~log.female.wt+log.neonatal.wt,
    data = perisso$trait %>% .[rownames(.) != "Tapirus pinchaque",],
    pem = pem.train
  ) -> lm1
  
  lm1$forward()
  ## summary(lm1)
  ## anova(lm1)
  
  predict(
    object = lm1,
    newdata = perisso$trait["Tapirus pinchaque",],
    newloc = train$location
  ) -> prd
  
  perisso$trait["Tapirus pinchaque","log.gestation.length"] <- prd
  
  ## Cross-validation:
  data.frame(
    observed = perisso$trait$log.neonatal.wt,
    auxiliary = NA,
    predictions = NA,
    lower = NA,
    upper = NA,
    row.names = rownames(perisso$trait)
  ) -> perisso.pred
  
  model.data(
    formula = log.neonatal.wt~log.female.wt+log.gestation.length,
    data = perisso$trait
  ) -> mdat
  
  ## i=1L
  for(i in 1L:nrow(perisso.pred)) {
    
    locate(
      x = net[[k]],
      target = rownames(perisso.pred)[i]
    ) -> train
    
    pem.train <- PEM(train$x, a = 0)
    
    evolution.model(
      object = pem.train,
      y = mdat$y[-i],
      x = mdat$x[-i,]
    ) -> opt
    
    pemlm(
      formula = log.neonatal.wt~log.female.wt+log.gestation.length,
      data = perisso$trait[-i,],
      pem = pem.train
    ) -> lm_cv
    
    predict(
      lm_cv$auxModel,
      perisso$trait[i,]
    ) -> perisso.pred[i,2L]
    
    lm_cv$forward()
    
    predict(
      object = lm_cv,
      newdata = perisso$trait[i,],
      newloc = train$location,
      interval = "prediction"
    ) -> perisso.pred[i,3L:5L]
  }
  
  perisso.pred
}

expl <- list()
psq <- c()
for(i in 1L:length(net)) {
  expl[[names(net)[i]]] <- runExample(i)
  expl[[i]] %>%
    {Psquare(.$observed, .$predictions)} -> psq[names(net)[i]]
}
```

```{r echo=FALSE}
figCounter$register(
  "psquare",
  paste(
    "Observed and predicted values of neonatal body mass for 13 odd-toed",
    "ungulate species across 11 phylogenetic models, with corresponding",
    "prediction coefficients shown in the lower right corner. Neonatal body",
    "mass was predicted using female body mass, gestation period, and",
    "phylogenetic eigenvectors derived from each of the phylogenetic graphs",
    "(labeled Net0--Net10). The best predictions are obtained from Net5, the",
    "model derived after adding five reticulations to the initial tree."
  )
)
```

The observed log-transformed gestation length for the mountain tapir is $2.61$
($\approx \,406\,\mathrm{days}$). Applying the same predictive workflow to the initial tree (Net0) and each of the ten reticulated networks (Net1--Net10)
enables us to evaluate how the inclusion of reticulation events influences trait
prediction accuracy (Figure  `r figCounter$getNumber("psquare")`).

---

```{r, echo=FALSE, fig.width=10, fig.height=7}
par(mfrow=c(3,4), mar=c(5.1,5.1,0.1,0.1))

rng <- sapply(expl, function(x) range(10^(x - 3)))
rng <- c(min(rng[1L,]),max(rng[2L,]))

log2kg <- function(x) 10^(x - 3)

## i="Net0"
for(i in names(net)) {
  plot(
    NULL, xlim = rng, ylim = rng, xlab = "Observed (kg)",
    ylab = "Predicted (kg)", asp=1, log="xy", las=2
  )
  arrows(x0=log2kg(expl[[i]]$observed), x1=log2kg(expl[[i]]$observed),
         y0=log2kg(expl[[i]]$lower), y1=log2kg(expl[[i]]$upper), angle=90,
         code=3, length=0.025)
  points(x=log2kg(expl[[i]]$observed), y=log2kg(expl[[i]]$predictions),
         pch=21L, bg="black")
  text(x=3, y=200, labels=parse(text=sprintf("bold(%s)",i)), adj=0, cex=1.25)
  text(
    x = 3,
    y = 100,
    labels = parse(text=sprintf("paste(italic(P)^2,\" = \", %0.3f)",psq[i])),
    adj = 0,
    cex=1.25
  )
  abline(0,1)
}
## main = sprintf("%s (%0.3f)", i, psq[i])


par(mar=c(5.1,5.1,1,0.1))
barplot(psq, names.arg=names(net), las=2, ylim=c(0,1.05), space = 0.1,
        ylab=expression(paste("Prediction coefficient (",italic(P^2),")")))
wh <- which.max(psq)
text(x=1.1*(wh - 0.5), y=psq[wh] + 0.075, label="*", cex=3, xpd=FALSE)
```

```{r, echo=FALSE, results='asis'}
figCounter$getCaption("psquare")
```

---


## Results

The prediction coefficients ranged from `r round(min(psq),3)` to
`r round(max(psq),3)`, with the best predictions obtained from the PEM
eigenvectors associated with the five-reticulation graph (Net5) and the worst
predictions obtained using the initial tree (Net0).


## Conclusion

This vignette demonstrates the application of Phylogenetic Eigenvector Maps
(PEMs) within the RPEM package to model and predict trait evolution in a
real-world phylogenetic context—specifically, among odd-toed ungulates
(Perissodactyla). By reconstructing an updated phylogeny from mitochondrial DNA
sequences retrieved from GenBank and progressively incorporating reticulation
events into the evolutionary model, we illustrate how accounting for
non-tree-like processes can improve the accuracy of comparative analyses.

Our results show that trait prediction performance, measured by the
cross-validated prediction coefficient $P^2$, varies significantly across
different phylogenetic structures. The highest predictive accuracy was achieved
using a reticulated network with five lateral evolutionary events (Net5),
outperforming both the original tree and other network configurations. This
suggests that, in this system, modeling gene flow or hybridization---even if
only implicitly through added reticulations—leads to more biologically realistic
representations of evolutionary history and enhances downstream inference.

Importantly, the workflow implemented here---from sequence-based phylogeny
reconstruction, through network augmentation, to PEM-based
prediction---highlights the flexibility and integrative power of the RPEM
framework. It enables users to move beyond strictly bifurcating trees and
explore the impact of reticulate evolution on trait dynamics, all within a
statistically rigorous and computationally efficient environment.

The dataset and code presented here serve as a reproducible example for
researchers interested in integrating genomic data, phylogenetic networks, and
trait modeling. We encourage further exploration of reticulate evolutionary
models and their implications for ecological and evolutionary inference using
the methods and software described.


## References
