---
title: "Phylogenetic Modelling and Prediction Using Phylogenetic Eigenvector Maps (PEM) in RPEM"
author: "Guillaume Guénard"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
bibliography: ../inst/bib/RPEM_tutorial.bib
vignette: >
  %\VignetteIndexEntry{Phylogenetic Modelling and Prediction Using Phylogenetic Eigenvector Maps (PEM) in RPEM}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})
##
### Load packages here:
##
### Figure counter:
(
  function() {
    log <- list(
      labels = character(),
      captions = character()
    )
    list(
      register = function(label, caption) {
        log$labels <<- c(log$labels, label)
        log$captions <<- c(log$captions, caption)
        invisible(NULL)
      },
      getNumber = function(label) {
        which(log$labels == label)
      },
      getCaption = function(label) {
        a <- which(log$labels == label)
        cap <- log$captions[a]
        cat(sprintf("Figure %d. %s\n\n---\n",a,cap))
        invisible(NULL)
      }
    )
  }
)() -> figCounter
```

Package version: `r packageVersion("RPEM")`


## Introduction

Phylogenetic Eigenvector Maps (PEM) provide a framework for phylogenetic
modelling--statistical approaches that incorporate phylogenetic relationships to
model trait evolution and predict trait values across species. These models are
particularly valuable when trait data are missing, difficult to measure, or
unavailable for certain species, such as rare, endangered, or extinct taxa
[@Guenard2013; @Guenard2011; @Guenard2014]. By leveraging shared evolutionary
history, PEM enables robust predictions even when empirical data are limited.

To apply phylogenetic modelling, two key inputs are required:

1. A set of species with known phylogeny and measured trait values (hereafter
   referred to as *model species*), and

2. The phylogenetic positions of the species for which traits are to be
   predicted (hereafter *target species*), ideally placed within or relative to
   the same phylogeny.

Moreover, phylogenetic modelling can be extended to incorporate functional or
ecological traits that are correlated with the focal trait. When such auxiliary
traits are available (or can be estimated) for target species, they can be
included as additional predictors to improve prediction accuracy--a process
known as *trait correlation modelling*.

While phylogenetic trees are traditionally described using biological terms like
*branches* and *tips*, this vignette adopts terminology from graph theory for
greater mathematical precision. Accordingly, we use *edge* instead of *branch*,
and *vertex* instead of *node*, *root*, or *tip*--except where specific
biological meaning must be preserved. This convention aligns with the formal
underpinnings of PEM and facilitates clearer algorithmic descriptions in the
RPEM package.

## The Phylogenetic Eigenvector Maps (PEM) Framework

The term *statistical map* traditionally refers to a thematic geographic map
that visualizes the spatial distribution of a variable, such as population
density or disease incidence. In this context, [@Moran1948], in his paper "The
Interpretation of Statistical Maps", introduced early methods for assessing
spatial autocorrelation in such georeferenced data.

Building on these ideas, **Moran's Eigenvector Maps (MEM)**, a term introduced
by @Dray2006, refer to the eigenvectors derived from a spatial connectivity
matrix, where eigenvalues are proportional to **Moran’s I** statistic--a measure
of spatial autocorrelation [@Moran1950]. These eigenvectors, originally termed
**Principal Coordinates of Neighbor Matrices (PCNM)** by @Borcard2002, represent
spatial patterns at different scales and are used to model spatial structure in
ecological and genetic data.

**Phylogenetic Eigenvector Maps (PEM)** [@Diniz2012; @Guenard2013] extend this
eigenfunction-based approach to phylogenetic data. PEMs are sets of eigenvectors
derived from the phylogenetic distance or connectivity matrix, designed to
describe the hierarchical structure of a phylogenetic graph--whether a strictly
bifurcating (tree-like) or reticulated (network-like) topology. Each eigenvector
captures phylogenetic variation at a distinct evolutionary scale, from deep
divergences to recent splits.

A key distinction between MEM and PEM is that, in PEM, the eigenvalues are
**not** proportional to Moran’s I coefficients of the corresponding
eigenvectors. Instead, they reflect the spectral decomposition of the
phylogenetic structure itself, enabling the modeling of trait evolution and
phylogenetic signal without relying on spatial autocorrelation assumptions.

Thus, while MEM and PEM share a common mathematical framework
--eigendecomposition of a relational matrix--they are applied in different
domains: spatial ecology and phylogenetics, respectively.


### The PEM Workflow

The Phylogenetic Eigenvector Maps (PEM) workflow consists of four main steps:

1. **Compute the influence matrix** of the phylogenetic graph, which encodes the
   pairwise relationships among species based on phylogenetic connectivity.

2. **Specify a model of trait evolution** along the edges (branches) of the
   phylogenetic tree, such as Brownian motion or Ornstein-Uhlenbeck, to inform
   how traits are expected to diverge over evolutionary time.

3. **Calculate the left eigenvectors** of the weighted and centred influence
   matrix. These eigenvectors represent orthogonal patterns of phylogenetic
   variation across different evolutionary scales.

4. **Use the resulting eigenvectors as explanatory descriptors** in statistical
   models to predict trait values or analyze phylogenetic signal [@Guenard2013].

An R implementation of this approach is provided in the **RPEM** package. RPEM
was designed to make the PEM workflow as seamless and accessible as possible for
users across ecology, evolution, and comparative biology. 

The package is under active development, and user feedback is highly encouraged.
Suggestions for improvements, bug reports, and feature requests are welcome to
help make RPEM a more efficient, robust, and user-friendly tool for phylogenetic
modelling.


## Installation instructions

You can install the RPEM package from CRAN or directly from GitHub using the
**devtools** package as follows:

```{r, eval=FALSE}
## From the CRAN
install.packages("RPEM")

## Directly from GitHub:
install.packages("devtools")
devtools::install_github("guenardg/RPEM")
```

Once **RPEM** is installed, the first step to compute a PEM is to load the
package. RPEM depends on the packages **ape**, **magrittr**, and **sf**, which
are automatically loaded when you attach RPEM:

```{r load_package}
library(RPEM)
```

his will also load the required dependencies, enabling access to phylogenetic
data structures (**ape**), piping operators (**magrittr**), and
spatial/geometric data handling (**sf**), where applicable.


## Example Analysis

### Data preparation

For this tutorial, we will use the perissodactyla dataset from the **caper** R
package. This dataset, originally compiled by @Purvis1995, contains body size
and ecological trait data for members of the mammalian order Perissodactyla
(e.g., horses, rhinos, tapirs), along with a corresponding phylogenetic tree.

```{r load_data}
data(perissodactyla, package="caper")
```

```{r plot_phylogeny, echo=FALSE, fig.height=4, fig.width = 7}
par(mar=c(2,2,2,2))
plot(perissodactyla.tree)
par(mar=c(5,4,4,2))
figCounter$register(
  "theTree",
  paste(
    "The phylogenetic tree of Perissodactyla used in this example. This tree,",
    "derived from the `perissodactyla.tree` object in the **caper** package,",
    "includes 17 odd-toed ungulate species and serves as the basis for",
    "computing Phylogenetic Eigenvector Maps (PEMs). Tip labels correspond to",
    "species, and branch lengths represent evolutionary time or divergence."
  )
)
```
```{r, plot_phylogeny_cap, echo=FALSE, results='asis'}
figCounter$getCaption("theTree")
```

The perissodactyla dataset contains two main components: 

* `perissodactyla.tree`, a phylogenetic tree encompassing
  `r length(perissodactyla.tree$tip.label)` odd-toed ungulate species (Figure
  `r figCounter$getNumber("theTree")`), and

* `perissodactyla.data`, a data frame with trait information for these species.

The data frame includes four traits:

* $\log_{10}$​-transformed adult female body mass,

* $\log_{10}$​-transformed gestation period,

* $\log_{10}$​-transformed neonatal body mass, and

* a categorical trait indicating whether the species exhibits territorial
  behavior (yes/no).

These traits can be used to explore evolutionary patterns and to demonstrate
phylogenetic prediction using PEMs. The combination of a well-resolved phylogeny
and continuous (and one binary) trait(s) makes this dataset ideal for
illustrating the RPEM workflow.

```{r data_table, echo=FALSE, results="latex"}
knitr::kable(perissodactyla.data)
```

Four species have missing data for the territoriality trait: the Sumatran rhinoceros (*Dicerorhinus sumatrensis*), the Javan rhinoceros (*Rhinoceros sondaicus*), the mountain tapir (*Tapirus pinchaque*), and the South American tapir (*Tapirus terrestris*).

For the purposes of this tutorial, we conducted a brief literature and web-based search to impute these missing values. Based on available ecological information, we assigned the following values:

```{r data_completion}
## Impute missing territoriality data based on external sources:
perissodactyla.data[c(2,9,12,13),"Territoriality"] <- c("No","No","Yes","Yes")
```

> These imputed values are not the result of a formal scientific review or
> primary research and should not be used for inference or decision-making
> outside the context of this tutorial. Researchers interested in the behavioral
> traits of these species are encouraged to consult authoritative sources or
> subject-matter experts before drawing conclusions.

Before proceeding, it is essential to ensure that the species in the
phylogenetic tree match those in the trait data table, and that they are in the
same order. A quick inspection reveals that the data table contains information
for only `r nrow(perissodactyla.data)` of the
`r length(perissodactyla.tree$tip.label)` species present in the tree.

To align the tree with the data, we will match the tip labels of the tree to the
species names in the data table using their binomial (Latin) scientific names.
We create a character vector, `spmatch`, to store the corresponding data row
names (or species identifiers) for each tip in the tree. When no match is found,
the entry in `spmatch` is set to `NA`.

These `NA` values identify species in the tree that lack trait data and can
therefore be safely removed using the `drop.tip()` function from the **ape**
package:

```{r droping_species}
## Match tree tips to data species:
match(
  perissodactyla.tree$tip.label,
  perissodactyla.data[,1L]
) -> spmatch

## Drop tips not present in the data:
drop.tip(
  perissodactyla.tree,
  perissodactyla.tree$tip.label[is.na(spmatch)]
) -> perissodactyla.tree
```

Now that the tree and data contain the same species, we must ensure that the
species are ordered identically in both the phylogenetic tree and the trait data
table. Mismatches in ordering can lead to incorrect associations between species
and trait values, which would compromise downstream analyses.

```{r check_order}
cbind(perissodactyla.tree$tip.label, perissodactyla.data[,1L])
```

Since the species ordering between the pruned tree and the data table no longer
matches, we need to recalculate the `spmatch` vector using the reduced tree and
then reorder the trait data accordingly to ensure alignment:

```{r re-order_species}
## Recalculate spmatch for the pruned tree:
match(
  perissodactyla.tree$tip.label,
  perissodactyla.data[,1L]
) -> spmatch

## Reorder the data to match the tip labels of the cleaned tree:
perissodactyla.data[spmatch,] -> perissodactyla.data

## Verify that species now match exactly:
all(perissodactyla.tree$tip.label == perissodactyla.data[,1L])
```

As a final step in the data preparation process, we will replace the current row
names of the data frame with the binomial species names (to ensure consistency
with the tree) and remove the first column (which may contain redundant or
non-essential information, such as an index or identifier):

```{r change_rownames}
perissodactyla.data[,1L] -> rownames(perissodactyla.data)
perissodactyla.data[,-1L] -> perissodactyla.data
```

The prepared data now appear as follows:

```{r re-arranged_data, echo=FALSE, results="latex"}
knitr::kable(perissodactyla.data)
```

### Calculating **PEM**

#### Edge weighting function

As introduced earlier, we adopt the terminology of graph theory to describe the
phylogenetic tree: the tree is treated as a directed acyclic graph, where
branches are referred to as edges, and the root, internal nodes, and tips are
collectively called vertices.

A central feature of the PEM method is the ability to incorporate explicit
models of trait evolution along the edges of the phylogeny. This is achieved
through an edge weighting function, which transforms raw edge lengths (typically
in units of time or substitutions) into weights that reflect the expected
magnitude of trait change along each edge.

The default weighting function in RPEM is a power function defined as:

$$
w_{a,\psi}(\phi_{j})=
\begin{cases}
\psi\phi_j^{\frac{1-a}{2}} & \phi_{j}>0\\
0 & \phi_{j}=0,
\end{cases}
$$
where

* $\phi_j$ is the length of edge $j$,

* $a$ is the steepness parameter, controlling how rapidly trait divergence
  accumulates after speciation,
  
* $\psi$ is a scaling factor representing the overall rate of trait evolution.

This parameterization, introduced by @Guenard2013, allows flexible modeling of
evolutionary processes--from Brownian motion-like patterns ($a \approx 0$) to
punctuational or stabilizing scenarios--by adjusting $a$ and $\psi$.

```{r display_weighting, echo=FALSE, fig.height=5, fig.width = 7}
par(mar=c(4.5,4.5,1,7) + 0.1)
d <- seq(0, 2, length.out=1000)
a <- c(0,0.33,0.67,1,0.25,0.75,0)
psi <- c(1,1,1,1,0.65,0.65,0.4)
cc <- c(1,1,1,1,1,1,1)
ll <- c(1,2,2,2,3,3,3)
trial <- cbind(a, psi)
colnames(trial) <- c("a","psi")
ntrials <- nrow(trial)
nd <- length(d)
matrix(
  NA,
  ntrials,
  nd,
  dimnames=list(paste("a=", trial[,"a"], ", psi=", trial[,"psi"], sep=""),
                paste("d=", round(d,4), sep=""))
) -> w
for(i in 1:ntrials)
  w[i,] <- RPEM::PEMweights(d, trial[i,"a"], trial[i,"psi"])
plot(NA, xlim=c(0,2), ylim=c(0,1.6), ylab="Weight", xlab="Distance", axes=FALSE)
axis(1, at=seq(0,2,0.5), label=seq(0,2,0.5))
axis(2, las=1)
text(expression(paste(~~~a~~~~~~~psi)),x=2.2,y=1.57,xpd=TRUE,adj=0)
for(i in 1:ntrials) {
  lines(x=d, y=w[i,], col=cc[i], lty=ll[i])
  text(paste(sprintf("%.2f", trial[i,1]), sprintf("%.2f",trial[i,2]), sep="  "),
       x=rep(2.2,1), y=w[i,1000], xpd=TRUE, adj=0)
}
rm(d,a,psi,cc,ll,trial,ntrials,nd,w,i)
figCounter$register(
  "edgeWeighting",
  paste(
    "Output of the edge weighting function $w_{a,\\psi}(\\phi_j)$ for",
    "different values of the steepness parameter $a$ and evolution rate",
    "$\\psi$. Each curve shows how edge weights scale with phylogenetic",
    "distance ($\\phi_j$) under a specific parameter combination. Higher",
    "values of $a$ lead to moreabrupt increases in weight with distance,",
    "reflecting punctuational evolutionary models, while lower values",
    "(e.g., $a = 0$) correspond to gradual, neutral (Brownian-like) evolution.",
    " The scaling factor $\\psi$ adjusts the overall magnitude of the weights."
  )
)
```
```{r, display_weighting_cap, echo=FALSE, results='asis'}
figCounter$getCaption("edgeWeighting")
```

As the steepness parameter $a$ increases, the weight assigned to a given edge
rises more sharply with respect to phylogenetic distance (or evolutionary time;
Figure `r figCounter$getNumber("edgeWeighting")`). In the context of PEM, edge
weights represent the relative rate of trait evolution along that edge: higher
weights indicate greater expected trait change.

* When $a = 0$, trait evolution follows a neutral model, analogous to Brownian
  motion, where changes accumulate gradually and randomly along edges.

* When $a = 1$, edge weights become independent of edge length. This corresponds
  to a punctuational model of evolution, in which trait change occurs
  predominantly at speciation events (vertices): either abruptly under
  directional selection, or not at all under stabilizing selection.

Importantly, the parameters $a$  and $\psi$ need not be constant across the
entire phylogeny. **RPEM** allows for heterogeneous models of evolution, where
different clades or lineages may follow distinct evolutionary regimes. For
example, a trait may evolve neutrally at a constant rate along the deep branches
of the tree, but experience shifts in evolutionary dynamics--such as accelerated
change or stronger selection--along specific lineages leading to the tips.

The first step in constructing a PEM is to convert the phylogenetic tree into a
graph representation. This is done using the `as.graph()` function, which
transforms the tree into a format compatible with graph-theoretic operations:

```{r convert_to_graph}
gr1 <- as.graph(perissodactyla.tree)
```

Here is a snippet illustrating how the graph container used by **RPEM** stores
graph information:

```{r graph_storage,echo=FALSE}
str(gr1)
```

The graph object used by **RPEM** consists of two main components:

1. A **vertex data frame**, which contains information about the vertices (nodes
   and tips) of the phylogenetic tree.
   
2. An **edge data frame**, stored as an attribute named `edge`, which describes
   the connections between vertices.

The vertex data frame may include zero or more columns of vertex-specific data.
Its rows correspond to individual vertices, with row names providing unique
identifiers (typically the vertex labels). In the current example, `as.graph()`
adds a logical column, `$species`, indicating whether a given vertex represents
a tip (i.e., a species) or an internal node.

The `edge` attribute is a data frame where each row represents an edge. The
first two columns contain the indices of the **origin** and **destination**
vertices (in parent-to-child order), followed by optional columns for edge
properties. In this case, `as.graph()` includes a numeric column, `$distance`,
which stores the phylogenetic edge lengths ($\phi_j$)--equivalent to the branch
lengths in the original `perissodactyla.tree`.

This structured representation enables flexible modeling of trait evolution
across the phylogeny, with both vertex and edge attributes accessible for
downstream computation in the PEM workflow.

Phylogenetic Eigenvector Maps are generated using the `PEM()` function. (In the
earlier package **MPSEM**, this functionality was provided by `PEM.build()`.)
The function takes, at minimum, a graph-class object—such as one created by
`as.graph()`--and returns a `PEM`-class object containing the computed 
eigenvectors and associated metadata.

By default, `PEM()` assumes a homogeneous model of trait evolution across the
entire phylogeny:

- The steepness parameter is set to $a = 0$, corresponding to a neutral model of
  evolution (analogous to Brownian motion), where trait change accumulates
  gradually along edges.
  
- The evolution rate is set to $\psi = 1$ uniformly across all edges.

This default setting implies that edge weights are proportional to the square
root of edge lengths ($w_j \propto \sqrt{\phi_j}$), under the power-law
weighting function described earlier.

The `PEM()` function is called as follows:

```{r PEM1}
## Calculate the PEM object using the default:
perissodactyla.PEM <- PEM(gr1)

## Show the PEM object:
perissodactyla.PEM
```

### Heterogeneous Evolutionary Processes

The **PEM** framework allows for modeling heterogeneous evolutionary processes
across the phylogenetic graph by permitting the parameters $a$ (steepness) and
$\psi$ (evolution rate) to vary across edges. This flexibility enables users to
define distinct evolutionary regimes in different clades or along specific
lineages.

Variation in $a$ and $\psi$ is modeled using **linear predictor submodels**
based on model matrices:

* One model matrix $\mathbf{M}_a$ for the steepness parameter $a$,

* One model matrix $\mathbf{M}_\psi$ for the evolution rate $\psi$.

Each model matrix has:

* One row per edge in the graph,

* One column per parameter in the submodel.

These matrices are multiplied by corresponding parameter vectors $\mathbf{b}_a$
and $\mathbf{b}_\psi$ to form linear predictors
$\eta_i = \mathbf{M}_a \mathbf{b}_a$ or
$\eta_i = \mathbf{M}_\psi \mathbf{b}_\psi$, which are then transformed via link
functions to constrain parameter values within biologically meaningful ranges.

#### Parameter $a$: Constrained via Logistic Transformation

The value of $a_i$ for edge $i$ is computed as:

$$
a_i = \frac{1}{1 + e^{-\eta_i}},
$$

which ensures $a_i \in (0, 1)$. This logistic transformation allows smooth
interpolation between near-neutral evolution ($a_i \to 0$) and punctuational
change ($a_i \to 1$).

By default:

* $\mathbf{M}_a$ is a single column of ones,

* $\mathbf{b}_a = -\infty$,

resulting in $a_i = 0$ for all edges—corresponding to a neutral (Brownian
motion-like) model of trait evolution.

#### Parameter $\psi$: Constrained Evolution Rate

The value of $\psi_i$ for edge $i$ is given by:

$$
\psi_i = \frac{2}{1 + e^{-\eta_i}},
$$

ensuring $\psi_i \in (0, 2)$. This scaling allows for variable rates of
evolution across the tree.

By default:

* $\mathbf{M}_\psi$ is a **zero-column matrix**,

* $\mathbf{b}_\psi$ is a **zero-length vector**,

which implies $\eta_i = 0$ for all edges, and thus $\psi_i = 1$. This
corresponds to a uniform evolution rate across the phylogeny.

### Defining Model Matrices

The entries in $\mathbf{M}_a$ and $\mathbf{M}_\psi$ are **binary indicators**
(0 or 1), used to assign parameter effects to specific subsets of edges. These can be constructed manually or using helper functions:

* `isUnderVertex(graph, v)` returns a binary vector with `1` for all edges
  **descending from vertex `v`** (i.e., in the subtree rooted at `v`), and `0`
  elsewhere.

* `isUnderEdge(graph, e)` returns a binary vector with `1` for all edges
  **descending from the destination vertex of edge `e`**, and `0` otherwise.

These functions are particularly useful for defining clade-specific evolutionary
models--for example, allowing accelerated evolution in a particular lineage.

The following figure illustrates the edge indexing and hierarchical structure of
the graph, helping identify which edges are "under" a given vertex or edge:

```{r tree_labelled, fig.height=5, fig.width = 7}
perissodactyla.tree -> tree
sprintf("N%s",1L:tree$Nnode) -> tree$node.label

par(mar=c(2,2,2,2))
plot(tree, show.node.label=TRUE)

edgelabels(
  sprintf("E%d",1L:nrow(tree$edge)),
  edge=1L:nrow(tree$edge),
  bg="white",
  cex=0.75
)
```
```{r tree_labelled_cap, echo=FALSE, results='asis'}
figCounter$register(
  "trainingTree",
  paste(
    "The labelled phylogenetic tree of the training species used in this",
    "example. This tree, derived from `perissodactyla.tree`, includes species",
    "labels at the tips and is used to compute Phylogenetic Eigenvector Maps",
    "(PEMs). Internal nodes and edges are implicitly represented, with branch",
    "lengths preserved to reflect evolutionary divergence. This tree serves as",
    "the basis for model training and prediction in subsequent analyses."
  )
)
figCounter$getCaption("trainingTree")
```

For example, let us define a model matrix that allows different evolutionary
dynamics in specific clades. We will include:

* A contrast for all species in the genus *Equus*, which are descendants of
  vertex **N9**,

* A second contrast for the entire lineage leading to the genus *Tapirus*, which
  branches off at edge **E11**.

This can be achieved by constructing a model matrix using the helper functions
`isUnderVertex()` and `isUnderEdge()`, which generate binary indicator vectors
identifying edges under a given vertex or edge, respectively:

For instance, let us define a model matrix with a contrast for all species of
genus _Equus_, which are located under vertex **N9**, and a second contrast
vector for the complete lineage leading to species of genus _Tapirus_ from the
rest of the tree, which are all found under edge **E11** as follows:

```{r contrast_equus}
cbind(
  Equus_clade = isUnderVertex(gr1, "N9"),
  Tapirus_lineage = isUnderEdge(gr1,"E11")
) -> mm
mm
```

A **PEM** is generated on the basis of this model matrix using `PEM()` as
follows:

```{r calculate_PEM1}
## Calculate the PEM:
PEM(
  x = gr1,
  a = c(-6,5,8),            ## steepness sub model parameters
  psi = c(-1,1),            ## evolution rate sub model parameters
  mm_a = cbind(1, mm),      ## model matrix: steepness sub model
  mm_psi = mm               ## model matrix: evolution rate sub model
) -> perissodactyla.PEM

## Show the PEM object:
perissodactyla.PEM

## Extract the graph from within the PEM-class object:
gr <- perissodactyla.PEM$graph()

## Show the steepness and evolution rate at the edges:
round(edge(gr)[,c("a","psi")],3)
```

In this example, we assume distinct evolutionary regimes across different
clades, with the following parameter values:

* In the genus *Equus* (edges E15-E21):
  $a \approx 0.269$, $\psi \approx 0.538$,

* In the genus *Tapirus* (edges E9-E13):  
  $a \approx 0.881$, $\psi \approx 1.462$,

* In all other parts of the tree (edges E1-E8 and E14), including the path from
  the root to the common ancestor of *Equus* and *Tapirus*:
  $a \approx 0.002$, $\psi = 1$.

These values reflect varying modes of evolution: near-neutral change in most of
the tree, stronger punctuational tendencies in *Tapirus*, and intermediate
dynamics in *Equus*.

The model is implemented using binary indicator vectors in the model matrices
$\mathbf{M}_a$ and $\mathbf{M}_\psi$, allowing for clade-specific
parameterization. More complex designs--such as nested factors, interaction
terms, or continuous covariates--can also be incorporated, provided that
sufficient information is available to define or estimate the parameters.

This flexibility enables biologically informed modeling of heterogeneous
evolutionary processes across the phylogeny.

In addition to the phylogenetic graph, the `PEM()` function accepts the
evolution model matrices (`mm_a`, `mm_psi`), parameter vectors (`b_a`, `b_psi`),
and two optional arguments:

* `d`: the name of the edge attribute storing phylogenetic distances (e.g.,
  branch lengths). Default is `"distance"`.

* `sp`: the name of the vertex attribute indicating which vertices represent
  species (i.e., tips). Default is `"species"`.

These defaults correspond to the naming convention used by `as.graph()`, so they
can typically be omitted unless custom attribute names were used.

The resulting `pem`-class object is a complex structure designed to store all
components of the Phylogenetic Eigenvector Map, including eigenvectors,
eigenvalues, input parameters, and metadata. While we will not explore its
internal structure in detail here, **RPEM** provides convenient extractor
methods to access the key output.

Using `as.matrix()` or `as.data.frame()`, you can extract the PEM eigenvectors
as a matrix or data frame, respectively. For a phylogeny with $n$ species, at
most $n - 1$ orthogonal eigenvectors are returned—each representing a distinct
mode of phylogenetic variation across the tree.

Below are the phylogenetic patterns of variation described by the first two
eigenvectors from the PEM computed earlier:

```{r Eigenvector_example, fig.height=4, fig.width=7}
tmp <- par(no.readonly = TRUE)
par(mfrow=c(1,2), mar=c(1.1,1.1,2.6,0.1))

## Singular vectors are extracted using the as.matrix method:
perissodactyla.U <- as.matrix(perissodactyla.PEM)

plot(perissodactyla.tree, x.lim=60, cex=1.0)

par(mar=c(1.1,0.1,2.6,1.1))
plot(NA, xlim=c(1,ncol(perissodactyla.U)), ylim=c(1,nrow(perissodactyla.U)),
     ylab="", xlab="", axes=FALSE, cex=1.5)

axis(3, at=1:ncol(perissodactyla.U), tick=FALSE, cex.axis=1.1,
     label=parse(text=sprintf("bold(u)[%d]",1:ncol(perissodactyla.U))))

absrng <- max(abs(perissodactyla.U))

for(i in 1:ncol(perissodactyla.U))
  points(
    x = rep(i,nrow(perissodactyla.U)),
    y = 1:nrow(perissodactyla.U),
    cex = 4*abs(perissodactyla.U[,i])/absrng,
    bg = grey(c(0,1))[1.5 + 0.5*sign(perissodactyla.U[,i])],
    pch=22
  )

par(tmp)
```
```{r Eigenvector_example_cap, echo=FALSE, results='asis'}
figCounter$register(
  "eigenvectorExample",
  paste(
    "A set of 12 eigenvectors derived from the perissodactyla phylogeny.",
    "Marker size scales linearly with the absolute value of the eigenvector",
    "loadings at each tip, while background color indicates the sign of the",
    "loading (white: negative, black: positive). Each panel represents one",
    "eigenvector, visualized across the species in the tree, illustrating",
    "distinct phylogenetic patterns of variation captured by the Phylogenetic",
    "Eigenvector Maps (PEM)."
  )
)
figCounter$getCaption("eigenvectorExample")
```

For instance, the first eigenvector primarily contrasts the Equids (horses and
relatives) with all other odd-toed ungulates, capturing the deepest phylogenetic
split in the Perissodactyla. The second eigenvector mainly differentiates tapirs
(genus *Tapirus*) from rhinocerotids (rhinoceros species), reflecting a major
clade divergence within the order. The remaining 10 eigenvectors represent
progressively finer-scale phylogenetic structures, each describing distinct
patterns of evolutionary relatedness at lower taxonomic levels.

### Estimating Weighting Parameters Empirically

The `PEM` class provides the method `evolution.model()` to empirically estimate
the optimal parameters of the edge weighting function--$a$ and $\psi$--which are
often unknown a priori.

Starting from the initial parameter values specified when creating the
`PEM`-class object, `evolution.model()` uses
**restricted maximum likelihood (REML)** to estimate the best-fitting
evolutionary model based on one or more response traits provided via the `y`
argument. Optionally, users can include auxiliary predictor traits through the
`x` argument, allowing the model to account for trait correlations when
estimating phylogenetic signal and evolutionary parameters.

This approach enables data-driven inference of the evolutionary process
underlying the observed trait distribution, improving the biological realism and
predictive accuracy of the PEM framework.

The `evolution.model()` method requires a response variable--here, neonatal
weight--which is used to optimize the parameter vectors $\mathbf{b}_a$ and, if
applicable, $\mathbf{b}_\psi$. It is called as follows:

```{r PEM_opt1}
## A simpler, single-steepness, model:
perissodactyla.PEM <- PEM(gr1, a = 0)

evolution.model(
  object = perissodactyla.PEM,
  y = perissodactyla.data[,"log.neonatal.wt"]
) -> opt

opt
```

The `evolution.model()` function returns a list containing details about the
optimization process, including the estimated parameters, log-likelihood, and
convergence status.

Alternatively, the response trait (specified via the `y` argument) and any
auxiliary predictor traits (via the `x` argument)—such as adult female body mass
and territoriality in this example—can first be organized using the
`model.data()` function. This helper function ensures proper alignment with the
phylogeny and prepares the data in a format suitable for `evolution.model()`.
The call then proceeds as follows:

```{r PEM_opt2}
perissodactyla.PEM_aux <- PEM(gr1, a = 0)

## Data preparation:
model.data(
  formula = log.neonatal.wt~log.female.wt+Territoriality,
  data = perissodactyla.data
) -> mdat

evolution.model(
  object = perissodactyla.PEM_aux,
  y = mdat$y,
  x = mdat$x
) -> opt_aux

opt_aux
```

The `model.data()` function prepares and organizes the data for phylogenetic
modeling by constructing a list that includes:

* The response trait(s), stored in the element `$y`,

* Any auxiliary predictor traits, stored in the element `$x`.

The traits can be specified using a formula interface (e.g., `y ~ x1 + x2`) and
a data frame, similar to how `lm()` or `glm()` internally parse inputs before
model fitting. This ensures proper alignment of trait data with the phylogeny
and facilitates seamless integration into the `evolution.model()` workflow.

It is important to note that any call to `evolution.model()` automatically
updates the `PEM` object with the estimated evolutionary parameters.
Additionally, it computes the generalized least squares (GLS) estimate of the
response trait variance--or, when auxiliary traits are included, the residual
variance after accounting for those predictors--and stores it within the PEM 
object.

This variance estimate can later be accessed as follows:

```{r}
perissodactyla.PEM_aux$S2()
```

It is important to note that estimates of the steepness parameter (stored in the
`$optim$par` element of the **PEM** object)--and consequently, the resulting
phylogenetic eigenvectors--can vary depending on whether auxiliary traits are
included in the model.

When modeling the response trait alone, the evolutionary signal is estimated
from the raw trait variation. However, when one or more auxiliary traits are
included (via the `x` argument), the model estimates phylogenetic signal in the
*residuals* of the response trait after accounting for those predictors. This
can lead to different optimal values for the evolutionary parameters.

In the example above, for instance, the linear predictor for the steepness
parameter, $\eta_a$, was estimated as `r round(opt$par$a, 2)` in the first call
(without auxiliary traits), and as `r round(opt_aux$par$a, 2)` in the second
call (with adult female body mass included as an auxiliary trait). This
difference reflects how adjusting for correlated traits can shift the inferred
mode of evolution.

### Prediction Scores

In phylogenetic prediction, a target species may be positioned either at an
existing vertex (e.g., a tip or internal node) or along an edge--representing a
lineage diverging from its latest common ancestor (**LCA**) at a specific
evolutionary distance from the origin of that edge.

The `PEM` class provides a `predict()` method to compute PEM prediction scores
for any such location on the phylogenetic graph, whether at a vertex or at an
intermediate point along an edge. These scores represent the expected
contribution of each PEM eigenvector to the trait value at the target location,
enabling interpolation of trait values based on phylogenetic proximity.

Internally, `predict()` dispatches to one of two specialized functions:

* `$predictVertex()`: computes scores for locations at vertices,

* `$predictEdge()`: computes scores for locations along edges, given a distance
  from the edge's origin.

This allows seamless integration of species--extant, missing, or even hypothetical--into the phylogenetic framework for trait prediction.

The `predict()` method is called as follows:

```{r}
## Model without an auxiliary trait.
predict(
  object = perissodactyla.PEM,
  newdata = data.frame(
    row.names = sprintf("target_%d",1:6),
    ref = c(1,3,5,2,3,4),
    dist = c(NA,0.1,NA,NA,0.4,1.1),
    lca = c(0,2,2.3,0,2,1.1)
  )
)

## Model with two auxiliary traits.
predict(
  object = perissodactyla.PEM_aux,
  newdata = data.frame(
    row.names = sprintf("target_%d",1:6),
    ref = c(1,3,5,2,3,4),
    dist = c(NA,0.1,NA,NA,0.4,1.1),
    lca = c(0,2,2.3,0,2,1.1)
  )
)
```

The `newdata` argument is a three-column data frame specifying the phylogenetic
location of each target species. The columns are:

* `ref`: an integer vector indicating the index of the vertex or edge where the
  **LCA** with the model species is located,
  
* `dist`: a numeric vector specifying, for entries where `_ref_` refers to an
  edge, the distance from the **origin** of that edge to the LCA; for vertex
  references, this value should be `NA`,
  
* `lca`: a numeric vector giving the evolutionary distance from the LCA to the
  **target species**.

This structure allows targets to be placed either at existing vertices or along
any edge of the phylogenetic graph.

The `predict()` method returns a matrix of **PEM scores**--the predicted
loadings of each eigenvector at the target locations--which can be used as
phylogenetic predictors in regression models. The returned matrix has an
associated attribute `"vf"` (variance factor), which provides the
**increment in prediction variance** due to the distance between the LCA and the
target. This accounts for increasing uncertainty as predictions extend further
from known nodes.

### Phylogenetic Modelling

To model or predict trait values, PEM eigenvectors are used as explanatory 
descriptors (predictor variables) in statistical models. The method is
flexible--any suitable modeling approach can be employed, such as linear
regression, generalized least squares (GLS), or mixed-effects models.

The PEM eigenvectors, which capture phylogenetic patterns of variation at
different evolutionary scales, are extracted from the `PEM`-class object using
the `as.matrix()` or `as.data.frame()` methods:

```{r extract}
## Extract the eigenvector as a matrix:
as.matrix(perissodactyla.PEM)

## Extract the eigenvector as a data frame:
as.data.frame(perissodactyla.PEM)
```

Package **RPEM** includes a framework for ordinary least squares (OLS)
regression built upon R’s `lm()` function and its associated methods. This
framework is implemented in the function `pemlm()`, which simplifies the
integration of PEM eigenvectors into linear models. The function automatically
handles the extraction and inclusion of PEM eigenvectors as predictors, enabling
straightforward phylogenetic modeling.

`pemlm()` is called as follows:

```{r build_pemlm_models}
## A first pemlm model without an auxiliary trait:
pemlm(
  log.neonatal.wt~1,
  perissodactyla.data,
  perissodactyla.PEM
) -> lm1

## Summary of the first model:
summary(lm1)

## Anova of the first model:
anova(lm1)

## A second pemlm model with two auxiliary traits:
pemlm(
  log.neonatal.wt~log.female.wt+Territoriality,
  perissodactyla.data,
  perissodactyla.PEM_aux
) -> lm2

## Summary of the second model:
summary(lm2)

## Anova of the second model:
anova(lm2)
```

Models produced by `pemlm()` do not automatically include phylogenetic
eigenvectors as predictors. Instead, the selection and inclusion of PEM
eigenvectors are handled through an internal set of embedded methods, which
manage model building and variable selection. A full description of these
internal mechanisms is beyond the scope of this tutorial.

One key embedded function, `$forward()`, performs **forward selection** of PEM
eigenvectors based on the **corrected Akaike Information Criterion (AICc)**
[@Hurvich1993]. This stepwise procedure adds eigenvectors one at a time,
selecting at each step the vector that provides the greatest improvement in
model fit relative to complexity, until no further improvement is possible.

```{r forward_select_pemlm_models}
## Adding eigenfunctions to the first model until the smallest AICc is reached
lm1$forward()
lm1

## Calculating the summary and anova 
summary(lm1)
anova(lm1)

## Adding eigenfunctions to the second model until the smallest AICc is reached
lm2$forward()
lm2

summary(lm2)
anova(lm2)
```

### Making Predictions

The `perissodactyla` dataset contains a missing value for gestation length in
the mountain tapir (*Tapirus pinchaque*). Here, we demonstrate how to predict
this missing trait value using the **RPEM** framework.

To make a prediction, we first need to:

1. Construct a phylogenetic tree (graph) excluding *Tapirus pinchaque*, so that
   the species is treated as a target rather than part of the model.

2. Determine the phylogenetic location (i.e., latest common ancestor and
   distance) of *Tapirus pinchaque* relative to this reduced tree.

This process is streamlined using the `locate()` method, which identifies the
optimal placement of the target species on the reference phylogeny--either at a
vertex or along an edge--based on taxonomic or phylogenetic criteria:

```{r locate1}
sp <- "Tapirus pinchaque"     ## A selected species.
train <- locate(gr1, sp)      ## The locate method.
train
```

Here, we see that *Tapirus pinchaque* is located at a distance of
`r train$location$ladist` from its LCA, which itself is positioned at a distance
of `r round(train$location$dist, 4)` along edge `r train$location$ref`.

The `locate()` method returns a two-element list:

* `$x`: the modified phylogenetic graph (i.e., the residual tree) with the
  target species removed,

* `$location`: a three-column data frame specifying the phylogenetic placement
  of the target, with columns `ref` (edge or vertex index), `dist` (distance
  along the edge from its origin, or `NA` if at a vertex), and `lca` (distance
  from the LCA to the target species).

This structure allows for accurate prediction while accounting for phylogenetic
uncertainty and placement.

We now prepare the model data using all available traits for the species in the
residual tree:

```{r}
## Prepare the model data:
model.data(
  formula = log.gestation.length~log.female.wt+log.neonatal.wt+Territoriality,
  data = perissodactyla.data
) -> mdat
```

To make predictions, we need to build a PEM using the residual graph returned by
the `locate()` method, as follows:

```{r}
## Initial PEM (single a = 0.5, single psi = 1) built from the residual graph:
pem.train <- PEM(train$x, a = 0)

## Estimate the evolution model:
evolution.model(
  object = pem.train,
  y = mdat$y[names(mdat$y) != sp],
  x = mdat$x[rownames(mdat$x) != sp,]
) -> opt

## Build a linear model:
pemlm(
  formula = log.gestation.length~log.female.wt+log.neonatal.wt+Territoriality,
  data = perissodactyla.data %>% .[rownames(.) != sp,],
  pem = pem.train
) -> lm3

## Summary and anova for the auxiliary trait model:
summary(lm3)
anova(lm3)

## Add the phylogenetic eigenfunctions:
lm3$forward()

## Summary and anova for the final model:
summary(lm3)
anova(lm3)
```

Finally, we use the `predict()` method--available for `pemlm`-class objects--to
obtain the predicted trait value as follows:

```{r}
## Make the prediction:
predict(
  object = lm3,
  newdata = perissodactyla.data[sp,],
  newloc = train$location
) -> prd
prd

## Substitute the missing value for the estimated gestation time:
perissodactyla.data[sp,"log.gestation.length"] <- prd
```

### Cross-validation of PEM Predictions

This section demonstrates how to perform leave-one-out cross-validation (LOOCV)
of trait predictions using the **PEM** framework and the R code presented in
previous sections. This approach assesses the predictive accuracy of the model
by iteratively removing one species, predicting its trait value based on the
remaining phylogeny and data, and comparing the prediction to the observed
value.

Predictions will be stored in a data frame called `perissodactyla.pred`, which
will include observed values, predicted values, and prediction errors. The
cross-validation is carried out as follows:

```{r cross-validation}
## Model formula:
f <- log.neonatal.wt~log.female.wt+log.gestation.length+Territoriality

## Table storing the results:
data.frame(
  observed = perissodactyla.data$log.neonatal.wt,
  auxiliary = NA,
  predictions = NA,
  lower = NA,
  upper = NA,
  row.names = rownames(perissodactyla.data)
) -> perissodactyla.pred

## Obtaining the updated model data:
model.data(
  formula = f,
  data = perissodactyla.data
) -> mdat

## For each species i:
for(i in 1L:nrow(perissodactyla.pred)) {
  
  ## Calculate the residual graph and location:
  train <- locate(gr1, rownames(perissodactyla.pred)[i])
  
  ## Calculate a PEM:
  pem.train <- PEM(train$x, a = 0)
  
  ## Estimate the evolution model:
  evolution.model(
    object = pem.train,
    y = mdat$y[-i],
    x = mdat$x[-i,]
  ) -> opt
  
  ## Fit the base model (auxiliary trait only):
  pemlm(
    formula = f,
    data = perissodactyla.data[-i,],
    pem = pem.train
  ) -> lm_cv
  
  ## Make prediction using the base model:
  predict(
    lm_cv$auxModel,
    perissodactyla.data[i,]
  ) -> perissodactyla.pred[i,2L]
  
  ## Add the PEM eigenfunction(s) on the basis of the AICc:
  lm_cv$forward()
  
  ## Make the prediction using the PEM-based model, including the limits of the
  ## 95% prediction interval:
  predict(
    object = lm_cv,
    newdata = perissodactyla.data[i,],
    newloc = train$location,
    interval = "prediction"
  ) -> perissodactyla.pred[i,3L:5L]
}

## Prediction coefficient using the auxiliary traits alone:
Psquare(perissodactyla.pred$observed, perissodactyla.pred$auxiliary)

## Prediction coefficient using the PEM eigenfunctions:
Psquare(perissodactyla.pred$observed, perissodactyla.pred$predictions)
```

```{r plot_pred_obs, echo=FALSE, fig.height=7, fig.width=7}
## Calculate the range of the whole data (predictions and interval limits):
rng <- range(perissodactyla.pred)

## Save the graphical parameters:
p <- par(no.readonly = TRUE)

## Generates an empty plot:
par(mar=c(5,5,2,2))
plot(NA, xlim=rng, ylim=rng, xlab="Observed (log) neonatal",
     ylab="predicted (log) neonatal", las=1, asp=1)

## Show the predictions without the PEM:
points(x=perissodactyla.pred$observed, y=perissodactyla.pred$auxiliary, pch=21,
       bg="blue")

## Show the predictions and their prediction intervals with the PEM:
arrows(x0=perissodactyla.pred$observed, x1=perissodactyla.pred$observed,
       y0=perissodactyla.pred$lower, y1=perissodactyla.pred$upper, code=3,
       angle=90, length=0.05)
points(x=perissodactyla.pred$observed, y=perissodactyla.pred$predictions,
       pch=21, bg="black")

## 1:1 line:
abline(0,1)

## Restore the graphical parameters:
par(p)

figCounter$register(
  "crossPreds",
  paste(
    "Leave-one-out cross-validated prediction of $\\log_{10}$​-transformed",
    "neonatal body mass for", nrow(perissodactyla.data), "odd-toed ungulate",
    "species. Each point represents a species, with predicted values obtained",
    "using Phylogenetic Eigenvector Maps (PEM) based on the residual phylogeny",
    "and trait data. The dashed 1:1 line indicates perfect prediction.",
    "Deviations from the line reflect prediction error, with the overall",
    "accuracy summarized by the cross-validated $R^2$. This plot demonstrates",
    "the predictive performance of the PEM framework in reconstructing missing",
    "or unobserved trait values in a phylogenetic context."
  )
)
```
```{r plot_pred_obs_cap, echo=FALSE, results='asis'}
figCounter$getCaption("crossPreds")
```

From the present cross-validation, we found that $\log_{10}$-transformed
neonatal body mass can be predicted with a cross-validated $R^2$ of
`r round(Psquare(perissodactyla.pred$observed, perissodactyla.pred$predictions),2)` (Fig.
`r figCounter$getNumber("crossPreds")`).

### Reticulated Graphs

Phylogenetic Eigenvector Maps (PEM) are not limited to strictly bifurcating
trees; the method is also applicable to reticulated phylogenetic graphs, such as
those representing hybridization, horizontal gene transfer, or network-based
evolutionary relationships. This flexibility makes PEM a powerful tool for
modeling trait evolution in clades with complex evolutionary histories.

To illustrate this capability, we will convert the original phylogenetic tree
from our previous example into a reticulated phylogenetic graph by adding one or
more reticulation edges (e.g., hybridization events). This extended graph
structure can then be processed using the same `as.graph()` and `PEM()`
functions, allowing eigenvector decomposition and phylogenetic modeling to
proceed analogously to the tree-based case.

```{r, echo=FALSE, fig.height=7, fig.width=7}
p <- par(no.readonly=TRUE)
par(mar=c(2,2,2,5))
plot(gr1, cex.min=4, cex.lab = 0.8, xpd=TRUE)
par(p)

figCounter$register(
  "graphPlotNodes",
  paste(
    "The original phylogenetic tree used in this example, with internal nodes",
    "labeled (e.g., N2, N8, etc.). These node labels are essential for",
    "specifying clade-specific evolutionary models in the Phylogenetic",
    "Eigenvector Maps (PEM) framework, particularly when using helper",
    "functions such as `isUnderVertex()` to define heterogeneous parameter",
    "regimes across branches. Tip labels correspond to species, and branch",
    "lengths reflect evolutionary divergence. This fully resolved tree serves",
    "as the basis for data alignment, model training, and phylogenetic",
    "prediction."
  )
)
```
```{r, echo=FALSE, results='asis'}
figCounter$getCaption("graphPlotNodes")
```

Let us assume there are three reticulations to be added to the phylogenetic
graph:

* An edge between vertices **N2** and **N10** with distance $d = 4$,

* A second edge between **N3** and **N7** with $d = 3$,

* A third edge between **N9** and **N7** with $d = 5$.

These reticulation edges are added using the `add.edge()` function as follows:

```{r}
## Add three edges to the original graph (tree):
gr1 %>%
  add.edge(
    from=c("N2","N3","N9"),
    to=c("N10","N7","N7"),
    data = data.frame(distance=c(5,7,5))
  ) -> gr_ret

## Show the new graph:
gr_ret
```

```{r, echo=FALSE, fig.height=7, fig.width=7}
p <- par(no.readonly=TRUE)
par(mar=c(2,2,2,5))
plot(gr_ret, cex.min=4, cex.lab = 0.8, xpd=TRUE)
par(p)

figCounter$register(
  "graphPlotRet",
  paste(
    "The phylogenetic tree from the example, transformed into a reticulated",
    "phylogenetic graph by adding three reticulation edges (e.g., representing",
    "hybridization or gene flow events). These edges connect nodes N2-N10,",
    "N3-N7, and N9-N7, introducing non-tree-like evolutionary relationships.",
    "This extended graph structure allows for modeling trait evolution in the",
    "presence of reticulate processes, demonstrating the flexibility of the",
    "Phylogenetic Eigenvector Maps (PEM) framework as implemented in RPEM."
  )
)
```
```{r, echo=FALSE, results='asis'}
figCounter$getCaption("graphPlotRet")
```

The rest of the analyses are essentially identical to the ones previously
carried out:

```{r}
## Build a PEM with single selection steepness and homogeneous evolution rate:
pem.ret <- PEM(gr_ret, a=0)

## Estimate the selection steepness (single):
evolution.model(
  object = pem.ret,
  y = mdat$y,
  x = mdat$x
) -> opt_ret

## Fit the base linear model:
pemlm(
  formula = log.neonatal.wt~log.female.wt+log.gestation.length+Territoriality,
  data = perissodactyla.data,
  pem = pem.ret
) -> lm_ret

## Summary and anova of the base model:
summary(lm_ret)
anova(lm_ret)

## Add the phylogenetic eigenfunctions:
lm_ret$forward()

## Summary an anova of the PEM model:
summary(lm_ret)
anova(lm_ret)
```

The $R^2$ values for both the baseline model and the PEM model are higher than
those obtained using the original tree--albeit by a small margin.
Cross-validation is performed in the same manner as before:

```{r}
## Table storing the results:
data.frame(
  observed = perissodactyla.data$log.neonatal.wt,
  auxiliary = NA,
  predictions = NA,
  lower = NA,
  upper = NA,
  row.names = rownames(perissodactyla.data)
) -> perissodactyla.pred_ret

## For each species i:
for(i in 1L:nrow(perissodactyla.pred)) {
  
  ## Calculate the residual graph and location:
  train <- locate(gr_ret, rownames(perissodactyla.pred)[i])
  
  ## Calculate a PEM:
  pem.train <- PEM(train$x, a = 0)
  
  ## Estimate the evolution model:
  evolution.model(
    object = pem.train,
    y = mdat$y[-i],
    x = mdat$x[-i,]
  ) -> opt
  
  ## Build an empty (auxiliary trait only) pemlm model:
  pemlm(
    formula = f,
    data = perissodactyla.data[-i,],
    pem = pem.train
  ) -> lm_cv
  
  ## Make prediction using the empty model:
  predict(
    lm_cv$auxModel,
    perissodactyla.data[i,]
  ) -> perissodactyla.pred_ret[i,2L]
  
  ## Add the PEM eigenfunction(s) on the basis of the AICc:
  lm_cv$forward()
  
  ## Make the prediction using the PEM-based model, including the limits of the
  ## 95% prediction interval:
  predict(
    object = lm_cv,
    newdata = perissodactyla.data[i,],
    newloc = train$location,
    interval = "prediction"
  ) -> perissodactyla.pred_ret[i,3L:5L]
}

## Prediction coefficient using the auxiliary traits alone:
Psquare(perissodactyla.pred_ret$observed, perissodactyla.pred_ret$auxiliary)

## Prediction coefficient using the PEM eigenfunctions:
Psquare(perissodactyla.pred_ret$observed, perissodactyla.pred_ret$predictions)
```

```{r, echo=FALSE, fig.height=7, fig.width=7}
## Calculate the range of the whole data (predictions and interval limits):
rng <- range(perissodactyla.pred_ret)

## Save the graphical parameters:
p <- par(no.readonly = TRUE)

## Generates an empty plot:
par(mar=c(5,5,2,2))
plot(NA, xlim=rng, ylim=rng, xlab="Observed (log) neonatal",
     ylab="predicted (log) neonatal", las=1, asp=1)

## Show the predictions without the PEM:
points(x=perissodactyla.pred_ret$observed, y=perissodactyla.pred_ret$auxiliary,
       pch=21, bg="blue")

## Show the predictions and their prediction intervals with the PEM:
arrows(x0=perissodactyla.pred_ret$observed, x1=perissodactyla.pred_ret$observed,
       y0=perissodactyla.pred_ret$lower, y1=perissodactyla.pred_ret$upper,
       code=3, angle=90, length=0.05)
points(x=perissodactyla.pred_ret$observed,
       y=perissodactyla.pred_ret$predictions, pch=21, bg="black")

## 1:1 line:
abline(0,1)

## Restore the graphical parameters:
par(p)

figCounter$register(
  "crossPreds_ret",
  paste(
    "Leave-one-out cross-validated prediction of log10​-transformed neonatal",
    "body mass for", nrow(perissodactyla.data), "odd-toed ungulate species,",
    "based on a reticulated phylogenetic graph. Predictions were generated",
    "using the Phylogenetic Eigenvector Maps (PEM) framework as implemented",
    "in RPEM, with the underlying graph incorporating three reticulation",
    "edges to model non-tree-like evolutionary processes (e.g., hybridization",
    "or gene flow). Each point represents a species, with values predicted",
    "from the remaining taxa after sequential exclusion. The dashed 1:1 line",
    "indicates perfect prediction. The cross-validated $R^2$ reflects the",
    "model's predictive accuracy, demonstrating that PEM maintains strong",
    "performance even when applied to reticulated phylogenies."
  )
)
```
```{r, echo=FALSE, results='asis'}
figCounter$getCaption("crossPreds_ret")
```

### Other Utility Functions

#### Influence Matrix

The **influence matrix** is a key component used internally by **RPEM** to
compute Phylogenetic Eigenvector Maps (PEMs). It encodes the topological
relationships between vertices (nodes and tips) and edges in the phylogenetic
graph.

This matrix has:

* One **row per vertex** (including both internal nodes and species tips),

* One **column per edge**.

Each element $I_{ij}$ is binary:

* $I_{ij} = 1$ if vertex $i$ is influenced by edge $j$,

* $I_{ij} = 0$ otherwise.

In this context, a vertex is said to be *influenced* by an edge if that edge
lies on the path from the vertex to the root—that is, if the edge contributes to
the evolutionary history leading to that vertex. Equivalently, this means the
vertex is a descendant of the edge’s destination.

The influence matrix can be extracted from a graph object using the
`influence()` function:

```{r influence_matrix}
InflMat(gr1) -> res
res
```

###1 Update a PEM with New Parameters

The calculation of the influence matrix--a core component in PEM computation--is
based solely on the topology of the phylogenetic graph and does not depend on
the edge weighting parameters. Therefore, when exploring different models of
trait evolution (e.g., varying $a$ or $\psi$), it is unnecessary to recompute
the influence matrix from scratch each time.

To improve computational efficiency, **RPEM** provides an `update()` method for
objects of class `PEM`. This method takes a previously computed `PEM` object,
applies new edge weighting parameters (via updated model matrices or parameter
vectors), and efficiently recalculates the eigenvectors without repeating the
full graph decomposition.

This allows for rapid exploration of evolutionary hypotheses while reusing the
underlying graph structure.

```{r PEM_updater}
## Update the PEM object, setting a to 0.9:
update(perissodactyla.PEM, a=logit(0.9), psi=NULL)

## Access the parameter values
perissodactyla.PEM$par()

## The parameter values are also available from the edge properties:
edge(perissodactyla.PEM$graph())
```

Any pre-existing estimate of the response trait variance is automatically
discarded when updating a PEM object. After updating the model parameters, the
variance must be re-estimated using the embedded method `$S2()` (to compute the
residual variance) or `$dev()` (to compute the deviance), based on the new
eigenvector decomposition and weighting scheme.

```{r forcedNewParameters}
## Old trait variance value:
perissodactyla.PEM$S2()

## Recalculating trait variance:
perissodactyla.PEM$S2(y = perissodactyla.data[,"log.neonatal.wt"])

## The new trait variance value is available thereafter:
perissodactyla.PEM$S2()
```

## Conclusion

The **RPEM** package provides a comprehensive and flexible framework for
phylogenetic modeling using Phylogenetic Eigenvector Maps. By integrating
graph-theoretic principles with statistical modeling, **RPEM** enables users to
explore trait evolution, predict missing trait values, and assess phylogenetic
signal across both tree-like and reticulated phylogenies.

Key features of the package include:

* Support for custom models of trait evolution via adjustable edge weighting,

* Tools for heterogeneous evolutionary processes across clades,

* Efficient prediction for species not included in the original model (via
  `locate()` and `predict()`),
  
* Built-in OLS modeling with `pemlm()`,

* Extensibility to phylogenetic networks (reticulated graphs).

While **RPEM** is still under active development, it already offers a robust
interface for phylogenetically informed modeling in ecology, evolution, and
comparative biology. Users are encouraged to provide feedback, report issues, or
contribute enhancements via the package repository.

We hope that **RPEM** will serve as a valuable tool for researchers seeking to
bridge phylogenetic structure with functional and ecological trait modeling.

## References
