---
title: "Simulating phylogenetic networks using RPEM"
author: "Guillaume Guénard"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
bibliography: ../inst/bib/Simulating_phylogenetic_network.bib
vignette: >
  %\VignetteIndexEntry{Simulating phylogenetic networks using RPEM}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})
##
### Load packages here:
##
### Figure counter:
(
  function() {
    log <- list(
      labels = character(),
      captions = character()
    )
    list(
      register = function(label, caption) {
        log$labels <<- c(log$labels, label)
        log$captions <<- c(log$captions, caption)
        invisible(NULL)
      },
      getNumber = function(label) {
        which(log$labels == label)
      },
      getCaption = function(label) {
        a <- which(log$labels == label)
        cap <- log$captions[a]
        cat(sprintf("Figure %d. %s\n\n---\n",a,cap))
        invisible(NULL)
      }
    )
  }
)() -> figCounter
```

Package version: `r packageVersion("RPEM")`


## Introduction

This vignette demonstrates the simulation capabilities of the **RPEM** package
for generating phylogenetic networks, DNA sequences, and trait data under
user-defined evolutionary models. These tools are designed to support method
development, hypothesis testing, and the creation of synthetic training
datasets--for example, in machine learning applications such as phylogenetic
graph reconstruction.

### Evolutionary Graphs in RPEM

The **RPEM** package includes functions to simulate evolutionary graphs, which
can represent a range of evolutionary scenarios:

* Linear sequences (e.g., simple ancestor-descendant chains),

* Bifurcating trees (i.e., phylogenies),

* Reticulated graphs (i.e., phylogenetic networks) that include hybridization or
  lateral gene transfer events.

Historically, species evolution has been modeled as a tree, under the assumption
that each species inherits genetic material from a single ancestor. This
framework supports only **vertical (longitudinal) gene transfer** and does not
account for **lateral gene transfer (LGT)** processes.

However, the tree model oversimplifies evolutionary history in many groups. It
ignores important biological phenomena such as:

* **Recombination**: the reshuffling and crossing-over of genetic material,

* **Hybridization**: reproductive events between distantly related lineages,

* **Horizontal exchange of mobile genetic elements**: e.g., plasmids,
  transposons, or viral insertions.

Representing evolution as a **reticulated graph**--rather than a strictly
bifurcating tree--allows for the explicit modeling of these lateral processes.
While this approach increases model complexity, it provides a more realistic
framework for studying clades where non-tree-like evolution plays a significant
role.

RPEM enables the simulation of such reticulated evolutionary histories, making
it a valuable tool for generating biologically plausible synthetic data under
complex evolutionary scenarios.

In this vignette, we demonstrate how to use **RPEM** to generate random
evolutionary graphs and simulate the evolution of DNA sequences and traits along
their edges. The resulting data--produced under known phylogenetic or network
structures and controlled evolutionary assumptions--can serve as benchmark
datasets for a variety of applications.

For example, such synthetic data are valuable for:

* Evaluating the performance of quantitative phylogenetic methods,

* Training and testing supervised and unsupervised machine learning models,

* Estimating evolutionary distances, inferring graph topology, or reconstructing
ancestral states.

By providing full control over the generative process, **RPEM** enables
rigorous, reproducible evaluation of phylogenetic inference tools and the
development of novel algorithms for evolutionary analysis.


## Simulating Evolutionary Graphs

Before proceeding, the **RPEM** package must be loaded into the R session using:

```{r}
library(RPEM)
```

### Molecular (DNA) Evolution Simulator 

RPEM includes a molecular evolution simulator that models DNA sequence evolution
along the edges of a phylogenetic graph as a continuous-time Markov process.
This process governs how nucleotides (A, G, C, T) change over evolutionary time
according to a specified substitution model.

The dynamics of this process are defined by a rate matrix Q, which describes the
instantaneous rates of change between nucleotide states:

$$
\mathbf{Q} = \begin{bmatrix}
-\Sigma_\mathrm{A} & r_\mathrm{AG} \pi_\mathrm{G} & r_\mathrm{AC} \pi_\mathrm{C} & r_\mathrm{AT} \pi_\mathrm{T} \\
r_\mathrm{GA} \pi_\mathrm{A} & -\Sigma_\mathrm{G} & r_\mathrm{GC} \pi_\mathrm{C} & r_\mathrm{GT} \pi_\mathrm{T} \\
r_\mathrm{CA} \pi_\mathrm{A} & r_\mathrm{CG} \pi_\mathrm{G} & -\Sigma_\mathrm{C} & r_\mathrm{CT} \pi_\mathrm{T} \\
r_\mathrm{TA} \pi_\mathrm{A} & r_\mathrm{TG} \pi_\mathrm{G} & r_\mathrm{TC} \pi_\mathrm{C} & -\Sigma_\mathrm{T}
\end{bmatrix},
$$

where:

* $r_{x,y}$ is the relative rate of substitution from nucleotide $x$ to $y$,

* $\pi_\mathrm{x}$ is the equilibrium frequency of nucleotide $x$,

* $\Sigma_x$ is the sum of the off-diagonal elements in row $x$, ensuring that
  each row of $\mathbf{Q}$ sums to zero (a requirement for valid Markov rate
  matrices).

For example:

$$
\Sigma_\mathrm{A} = r_\mathrm{AG} \pi_\mathrm{G} + r_\mathrm{AC} \pi_\mathrm{C} + r_\mathrm{AT} \pi_\mathrm{T}
$$

The model assumes: 

* The sum of equilibrium frequencies is

$$
\pi_\mathrm{A} + \pi_\mathrm{G} + \pi_\mathrm{C} + \pi_\mathrm{T} = 1
$$

* The 12 non-diagonal rate parameters $r_{x,y}$ (for $x \neq y$) are scaled so
that the average rate across all substitutions is 1.

These constraints ensure that the $\mathbf{Q}$ matrix is properly normalized,
allowing evolutionary time to be measured in expected substitutions per site.

In addition to nucleotide substitutions, RPEM simulates insertions and deletions
(indels) by extending the Markov process to include gaps (denoted as
${\text -}$) as an additional evolutionary state. This allows the model to
capture sequence length variation across lineages.

Gaps are treated as a fifth state in the evolutionary process: 

* A gap-to-nucleotide transition simulates an insertion,

* A nucleotide-to-gap transition simulates a deletion.

The full process is governed by an extended rate matrix $\mathbf{Q^*}$:

$$
\mathbf{Q^*} = \begin{bmatrix}
-\Sigma_\mathrm{\text -} & r_\mathrm{{\text -} N} (1 - \pi_{\text -}) \\
r_\mathrm{N {\text -}} \pi_{\text -} & \mathbf{Q} (1 - \pi_{\text -})
\end{bmatrix},
$$

where:

* $r_\mathrm{{\text -} N}$ is the insertion rate (gap $\rightarrow$ 
nucleotide),

* $r_\mathrm{N {\text -}}$ is the deletion rate (nucleotide $\rightarrow$ gap)
),

* $\pi_{\text -}$ is the equilibrium frequency of gaps in the alignment,

* $1 - \pi_{\text -}$  is the total equilibrium frequency of nucleotides,

* $\mathbf{Q}$ is the original nucleotide substitution rate matrix (as defined
  earlier),

* $\Sigma_\mathrm{\text -} = r_\mathrm{{\text -} N} (1 - \pi_{\text -})$ ensures
the row sums of $\mathbf{Q^*}$ are zero.

This block-structured formulation allows joint simulation of substitutions and
indels under a unified continuous-time Markov framework. The resulting sequences
include both base changes and alignment gaps, producing biologically realistic
alignments suitable for benchmarking phylogenetic inference methods or training
machine learning models.

The resulting $5 \times 5$ rate matrix $\mathbf{Q^*}$, which governs the joint
evolution of nucleotides (A, C, G, T) and gaps (${\text -}$), is defined as:

$$
\mathbf{Q^*} = \begin{bmatrix}
-\Sigma_\mathrm{{\text -}} & r_\mathrm{{\text -} A} \pi_\mathrm{A} (1 - \pi_{\text -}) & r_\mathrm{{\text -} G} \pi_\mathrm{G} (1 - \pi_{\text -}) & r_\mathrm{{\text -} C} \pi_\mathrm{C} (1 - \pi_{\text -}) & r_\mathrm{{\text -} T} \pi_\mathrm{T} (1 - \pi_{\text -}) \\
r_\mathrm{A {\text -}} \pi_{\text -} & -\Sigma_\mathrm{A} & r_\mathrm{AG} \pi_\mathrm{G} (1 - \pi_{\text -}) & r_\mathrm{AC} \pi_\mathrm{C} (1 - \pi_{\text -}) & r_\mathrm{AT} \pi_\mathrm{T} (1 - \pi_{\text -}) \\
r_\mathrm{G {\text -}} \pi_{\text -} & r_\mathrm{GA} \pi_\mathrm{A} (1 - \pi_{\text -}) & -\Sigma_\mathrm{G} & r_\mathrm{GC} \pi_\mathrm{C} (1 - \pi_{\text -}) & r_\mathrm{GT} \pi_\mathrm{T} (1 - \pi_{\text -}) \\
r_\mathrm{C {\text -}} \pi_{\text -} & r_\mathrm{CA} \pi_\mathrm{A} (1 - \pi_{\text -}) & r_\mathrm{CG} \pi_\mathrm{G} (1 - \pi_{\text -}) & -\Sigma_\mathrm{C} & r_\mathrm{CT} \pi_\mathrm{T} (1 - \pi_{\text -}) \\
r_\mathrm{T {\text -}} \pi_{\text -} & r_\mathrm{TA} \pi_\mathrm{A} (1 - \pi_{\text -}) & r_\mathrm{TG} \pi_\mathrm{G} (1 - \pi_{\text -}) & r_\mathrm{TC} \pi_\mathrm{C} (1 - \pi_{\text -}) & -\Sigma_\mathrm{T}
\end{bmatrix},
$$

where:

* $r_\mathrm{{\text -} x}$ is the insertion rate (gap $\rightarrow$ nucleotide
$x$),

* $r_\mathrm{x {\text -}}$ is the deletion rate (nucleotide $x$ $\rightarrow$ 
gap),

* $\pi_x$ is the equilibrium frequency of nucleotide $x$,

* $\pi_{\text -}$ is the equilibrium frequency of gaps,

* $1 - \pi_{\text -}$ is the total equilibrium probability of being in a
nucleotide state,

* $\Sigma_{\text -} = \sum_{x \in \lbrace \mathrm{A, G, C, T} \rbrace} r_{{\text -},x}\pi_x (1 - \pi_{\text -})$

* $\Sigma_x$ (for each nucleotide) includes both substitution and deletion
rates, ensuring that each row of $\mathbf{Q^*}$ sums to zero.

The probability transition matrix $\mathbf{P}$ is derived from the extended rate
matrix $\mathbf{Q^*}$ (which includes both nucleotide substitutions and indels)
using matrix exponentiation:

$$
\mathbf{P}(t) = e^{\rho t \mathbf{Q^*}}
$$

In this expression:

* $t$ represents the evolutionary time (in expected substitutions per site),

* $\mathbf{Q^*}$ is the instantaneous rate matrix (as defined previously),

* $\rho$ is the mean shift rate over all possible states,

* $e^{\rho t \mathbf{Q^*}}$  denotes the matrix exponential, which computes the
probabilities of transitioning from one state to another over time $t$,

$\mathbf{P}(t)$ is a $5 \times 5$ matrix where each element $P_{ij}(t)$ gives
the probability of starting in state $i$ and ending in state $j$ after time $t$.

Each row of $\mathbf{P}(t)$ sums to 1, as required for a valid probability
matrix.

RPEM implements eight commonly used models of DNA sequence evolution, all of
which are time-reversible (i.e., $r_\mathrm{xy} = r_\mathrm{yx}$ for all
nucleotide pairs $x$ and $y$). These include well-known models such as:

* Jukes–Cantor (JC69),

* Kimura 2-parameter (K80),

* Felsenstein 1981 (F81),

* Hasegawa–Kishino–Yano (HKY85),

* General Time Reversible (GTR), and others.

This functionality is provided by the `DNArate()` function, which generates the
corresponding rate matrix $\mathbf{Q^*}$ based on the selected model and
user-specified parameters (e.g., equilibrium frequencies,
transition/transversion rate ratios). The function is called as follows: 

```{r}
Q <- DNArate(model="JC69", piGap=0.3, insertionRate=0.1, deletionRate=0.1)
Q
```

This shift rate matrix corresponds to the JC69 model of DNA evolution
[@JC69model], assuming equal nucleotide frequencies
($\pi_\mathrm{A} = \pi_\mathrm{G} = \pi_\mathrm{C} = \pi_\mathrm{T} = 0.25$) and
time-reversible substitution rates. In this example, we extend the model to
include indels, with:

* An equilibrium gap frequency of $30\%$ ($\pi_{\text -} = 0.3$),

* Insertion and deletion rates set to 0.1 (relative to the substitution rate
  scale).

**RPEM** implements molecular evolution simulations through the function
`molEvolSim()`, which takes the following key arguments:

* The instantaneous rate matrix $\mathbf{Q^*}$ (including substitutions and
  indels),

* The time step $t$ (argument step), representing the length of each
  evolutionary branch or segment,

* The mean shift rate $\rho$ (argument `rho`), used to scale the overall rate of
  evolution (e.g., to model variation in evolutionary tempo across lineages).

A typical call to simulate DNA sequence evolution under this model is as
follows:

```{r}
em1 <- molEvolSim(Q, step = 1, rho = 5)
```

The `molEvolSim()` function returns a list containing three embedded functions
that allow flexible and efficient simulation of molecular evolution:

* `$recalculate(t, rho)` Updates the transition probability matrix
  $\mathbf{P}(t) = e^{\rho t \mathbf{Q^*}}$ for new values of evolutionary time
  $t$ (argument `t`) and mean shift rate $\rho$ (argument `rho`), without
  reconstructing the entire simulator.

* `$evolve(N)` Simulates the evolution of a nucleotide (or gap) over time. It
  takes an initial nucleotide state (as a raw character, e.g., "A" or "-") and
  returns a new state drawn from the posterior probability distribution after
  evolution along a branch of length $t$. This function can be called
  iteratively to simulate evolution across multiple steps.

* `$getMt()` Returns the current transition probability matrix $\mathbf{P}(t)$,
  computed from the rate matrix $\mathbf{Q^*}$ and the current values of $t$ and
  $\rho$.

This modular design allows users to simulate sequence evolution efficiently across complex phylogenetic graphs, including trees and reticulated networks.
For example: 

```{r}
em1$getMt()
```

returns the transition probability matrix $\mathbf{P}(t)$ for a time step of
$t = 1$  and a mean shift rate of $\rho = 5$. This matrix gives the
probabilities of evolving from one nucleotide (or gap) state to another over
that evolutionary duration. 

If we recalculate the simulator for a shorter time step, $t = 0.1$: 

```{r}
em1$recalculate(0.1, 5)
em1$getMt()
```

the transition probabilities become smaller, reflecting the reduced opportunity
for change over shorter evolutionary intervals. 

Conversely, if we increase the time step to $t = 5$:

```{r}
em1$recalculate(5, 5)
em1$getMt()
```

the transition probabilities increase, as more time allows for a higher
likelihood of substitution or indel events.

This dynamic behavior illustrates how the Markov process scales with time and
rate, enabling biologically realistic simulations across branches of varying
lengths. 

These tools provide fine-grained control over DNA sequence evolution, supporting
applications in: 

* Testing phylogenetic inference methods,

* Benchmarking sequence alignment algorithms,

* Generating training data for machine learning models in evolutionary genomics.

> In R, it is more efficient to work with raw vectors when manipulating
> individual characters of a character string. Unlike standard character
> strings, raw vectors allow direct access and modification of single-byte
> elements using standard subsetting operators (`[]`), avoiding the overhead of
> string-specific functions like `substr()` or `strsplit()`. After manipulation,
> the `rawToChar()` function can be used to convert the raw vector back into a
> character string. This approach is particularly useful in low-level text
> processing tasks, such as simulating DNA sequence evolution, where performance
> and byte-level control are important.

For example, let us simulate the evolution of a single DNA site starting with
the nucleotide "A" at time $t = 0$, evolving over 100 time steps, each of length
$t = 1$, under a specified substitution model (e.g., JC69) with mean shift rate
$\rho = 5$. This can be achieved using the molecular evolution simulator
function returned by `molEvolSim()`:

```{r}
set.seed(1324987)                 ## Set RNG seed for reproducibility.
em1$recalculate(1, 5)             ## Recalculate for a time step of 1
s <- raw(101)                     ## For storage
s[1] <- charToRaw("A")            ## Initialize with and "A"
for(i in 2:101)                   ## Simulation loop:
  s[i] <- em1$evolve(s[i - 1]) 
rawToChar(s)                      ## Show time series
```

This example shows multiple instances of nucleotide conservation (two or more
consecutive occurrences of the same nucleotide), substitutions (changes from one
nucleotide to another), deletions, and insertions--demonstrating the full range
of molecular evolutionary processes simulated by **RPEM**.

The assumption that all sites in a DNA sequence evolve at the same rate is often
violated in real biological data. In practice, sites exhibit rate heterogeneity,
with some evolving slowly (e.g., functionally constrained regions) and others
more rapidly (e.g., non-coding or loop regions).

To generate more biologically realistic simulations, RPEM allows users to assign
site-specific evolutionary rates drawn from a gamma distribution--a commonly
used model for among-site rate variation in phylogenetics. This distribution is
parameterized by a shape parameter $\alpha$, where lower values (e.g.,
$\alpha < 1$) reflect high rate variation, and higher values approach a constant
rate.

The function `drawEvolRate()` facilitates this process by generating random rate multipliers from a gamma distribution with user-defined shape and scale parameters:

```{r}
er <- drawEvolRate(NN=70, gamma.shape=3, gamma.scale=2e-2)
er
```

where: 

* `NN` is the number of sites in the DNA sequence,

* `gamma.shape` is the shape parameter of the gamma distribution, controlling
  the degree of rate variation across sites (lower values indicate greater
  heterogeneity),

* `gamma.scale` is the scale parameter, determining the spread of the
  distribution.

The resulting vector er contains site-specific evolutionary rate multipliers 
($\rho$), which can be applied during sequence simulation to scale the local
substitution rate along branches. This allows **RPEM** to generate more
biologically realistic DNA alignments, reflecting the well-documented phenomenon
of among-site rate variation observed in empirical data.

By incorporating rate heterogeneity, users can better emulate real-world evolutionary dynamics and improve the realism of benchmark datasets for phylogenetic inference, model testing, or machine learning applications. 


**RPEM** includes the function `drawDNASequence()` to generate random initial
DNA sequences for use in evolutionary simulations. This allows users to start
simulations from biologically plausible initial conditions, with control over
nucleotide and gap composition. 

The function is called as follows: 

```{r}
set.seed(13987463)    ## Set RNG seed for reproducibility.

## Generating a random DNA sequence
isq <- drawDNASequence(NN=length(er), piGap=0.2, pi=c(0.35,0.15,0.15,0.35))
rawToChar(isq)        ## Showing the random DNA sequence
```

where: 

* `NN` is the length of the sequence (here, matching the number of sites in
  `er`),

* `piGap` is the equilibrium frequency of gaps ($\pi_{\text -} = 0.2),

* `pi` is a vector of equilibrium nucleotide frequencies:
  $\pi_\mathrm{A} = 0.35$, $\pi_\mathrm{G} = 0.15$, $\pi_\mathrm{C} = 0.15$,
  $\pi_\mathrm{T} = 0.35$.

The resulting sequence is slightly AT-rich (since
$\pi_\mathrm{A} + \pi_\mathrm{T} = 0.7$) and includes a $20\%$ probability of
gaps, reflecting a realistic indel burden. Internally, the sequence is stored as
a raw vector for efficient manipulation during simulation, and can be converted
to a character string using `rawToChar()` when needed.

This functionality supports the generation of diverse, user-defined starting
points for molecular evolution simulations--enhancing the flexibility and
realism of synthetic data in phylogenetic method development.

We will now simulate the evolution of the initial DNA sequence under a realistic
model where all sites share the same underlying substitution and indel process
(defined by the shift rate matrix $\mathbf{Q^*}$), but each site evolves at its
own rate. This is achieved by assigning a site-specific mean shift rate
($\rho_j$), drawn earlier from a gamma distribution (e.g., using
`drawEvolRate()`).

To implement this, we create a list of molecular evolution simulators--one for
each site--where each simulator uses the same $\mathbf{Q^*}$ and time step, but
is scaled by its own $\rho_j$ value:

```{r}
nesl <- list()              ## A list of nucleotide evolution simulators

## Initialize each simulator with a different mean shift rate:
for(j in 1:length(er))
  nesl[[j]] <- molEvolSim(Q, step = 1, rho = er[j])
```

This approach allows site-specific rate variation, a well-documented feature of
molecular evolution: some sites (e.g., coding regions) evolve slowly, while
others (e.g., non-functional regions) accumulate changes more rapidly.

Each simulator in `nesl` can then be used independently to evolve its
corresponding site along a phylogenetic tree or network, enabling biologically
realistic sequence simulation that accounts for among-site rate heterogeneity--a
critical feature for generating benchmark datasets in phylogenetics and machine
learning.

To evolve the initial DNA sequence forward in time, we apply the site-specific
molecular evolution simulators to each nucleotide position. This is done by
calling the `$evolve()` method of each simulator in the list `nesl`, using the
corresponding site from the initial sequence as input:

```{r}
isq2 <- raw(length(isq))          ## Allocate storage for the evolved sequence
for(j in 1:length(isq))                   ## For each site j
  isq2[j] <- nesl[[j]]$evolve(isq[j])     ## Evolve nucleotide at site j
rawToChar(isq)                            ## Initial sequence
rawToChar(isq2)                           ## Evolved sequence
```

Each site evolves under the same substitution and indel process (defined by
$\mathbf{Q^*}$), but with a rate scaled by its own $\rho_j$ value drawn from a
gamma distribution. This results in biologically realistic variation in
evolutionary rates across sites.

The output shows how individual nucleotides (or gaps) change over a single time
step of length $t = 1$, reflecting the stochastic nature of molecular evolution.
Some sites remain conserved, while others undergo substitutions or
indels--demonstrating the combined effects of the underlying Markov process and
site-specific rate variation.

This modular simulation approach enables fine-grained control over sequence
evolution and supports the generation of complex, realistic datasets for
phylogenetic method evaluation, alignment benchmarking, or training machine
learning models in evolutionary genomics.

To simulate evolution over multiple time steps, we extend the process by adding
an outer loop that iterates across successive time intervals. At each step,
every site in the sequence is evolved according to its own site-specific
evolutionary rate, as defined by its corresponding simulator in the `nesl` list.

This two-level structure--sites in the inner loop, time steps in the outer
loop--allows for realistic, asynchronous evolution across the sequence:

```{r, fig.width = 7.0}
stbl <- matrix(raw(),51,length(er))           ## For storing 51 sequences
stbl[1,] <- isq                               ## Initial sequence
stbl[2,] <- isq2                              ## Second sequence

## Simulation loops from squence 3 to 51:
for(i in 3:nrow(stbl))                              ## Loop for each sequence i
  for(j in 1:ncol(stbl))                            ## Loop for each site j
    stbl[i,j] <- nesl[[j]]$evolve(stbl[i - 1,j])    ## Simulate the shifts

## Showing the DNA sequences:
stbl %>%
  concatenate %>%
  show.sequence
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "dnaSeq1",
  paste(
    "A set of 51 aligned DNA sequences generated by evolving a randomly",
    "generated ancestral sequence over 50 regular time steps. Nucleotides are",
    "color-coded as follows: red = Adenine (A), yellow = Guanine (G),",
    "green = Cytosine (C), blue = Thymine (T), and grey = gap (--). The",
    "sequences were simulated using a site-specific molecular evolution model",
    "with gamma-distributed rate variation across sites, as implemented in",
    "**RPEM**. This alignment represents a synthetic dataset suitable for",
    "benchmarking phylogenetic inference methods or training machine learning",
    "models in evolutionary genomics."
  )
)

figCounter$getCaption("dnaSeq1")
```

In this code example, we introduce two additional functions from RPEM: `concatenate()` and `show.sequence()`.

* The `concatenate()` function applies `rawToChar()` row by row to a raw matrix
  (e.g., a matrix of simulated sequences), converting each row into a character
  string. It optionally allows the removal of specified characters--most
  commonly gaps ("-")--during the conversion. This is useful for generating
  clean sequence alignments for export or analysis.

* The `show.sequence()` function is a graphical tool for visualizing sets of DNA
  sequences. It displays sequences as colored character matrices, with
  customizable color schemes for nucleotides (e.g., A = red, G = yellow,
  C = green, T = blue) and gaps (e.g., grey), enabling intuitive inspection of
  conservation, variation, and indel patterns.

For instance, one may display the simulated sequences while omitting gaps as
follows:

```{r, fig.width = 7.0}
## Display the DNA sequences without the gaps:
stbl %>%
  concatenate(discard="-") %>%
  show.sequence
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "dnaSeq2",
  paste(
    "A set of 51 DNA sequences generated by evolving a randomly generated",
    "ancestral sequence over 50 regular time steps, with gaps removed prior to",
    "visualization. Nucleotides are color-coded as follows: red = Adenine (A),",
    "yellow = Guanine (G), green = Cytosine (C), and blue = Thymine (T). The",
    "sequences were simulated using a site-specific molecular evolution model",
    "with gamma-distributed rate variation across sites, as implemented in",
    "**RPEM**. This alignment represents a synthetic dataset suitable for",
    "benchmarking phylogenetic inference methods or training machine learning",
    "models in evolutionary genomics."
  )
)

figCounter$getCaption("dnaSeq2")
```

This is how DNA sequences typically appear after sequencing: as aligned,
character-based strings of nucleotides. In this simulation, the sequences were a
priori aligned by construction--meaning that homologous sites evolved from a
common ancestral position--making this alignment "true" by design.

Such known alignments are invaluable for evaluating the accuracy and performance
of DNA sequence alignment algorithms and software (e.g., MAFFT, Clustal,
MUSCLE), as they provide a ground truth against which inferred alignments can be
compared.

When needed, the simulated sequences can be exported in standard FASTA format
using the RPEM function `write.fasta()`:

```{r}
stbl %>%
  concatenate(discard="-") %>%
  write.fasta(file="Test.fasta")
```

This FASTA file can be easily accessed and read by other software and R
packages, such as the **ape** package, which provides comprehensive tools for
phylogenetic analysis. 

For example, you can read the simulated sequences into R using the `read.dna()` function from **ape**:

```{r}
if(require(ape)) {
  tmp <- read.dna(file="Test.fasta", format="fasta", as.character=TRUE)
  tmp[1:5]
}
```

```{r, echo=FALSE, results='hide'}
file.remove("Test.fasta")
rm(tmp)
```

### Trait Evolution Simulator 

RPEM simulates quantitative trait evolution using the Ornstein-Uhlenbeck (OU)
process, a stochastic model that extends Brownian motion by incorporating
stabilizing selection toward an optimal trait value.

Unlike pure diffusion (Brownian motion), where trait values drift randomly over
time, the OU process introduces a pull toward an optimum ($x_{opt}$), simulating
evolutionary constraints such as stabilizing selection or adaptive peaks in a
fitness landscape. The strength of this pull is governed by the selection rate
$\alpha$: higher values of $\alpha$ result in tighter clustering around the
optimum, while lower values allow for greater divergence.

The trait value $x_t$ after evolutionary time $t$ is drawn from a normal
distribution:

$$
x_t \sim \mathcal{N} \left(x_0 e^{-\alpha t} + (1 - e^{-\alpha t}) x_{opt}, \sigma \sqrt{\frac{(1 - e^{-2\alpha t})}{2 \alpha}} \right),
$$

where:

* $x_0$ : initial trait value at time 0,

* $x_t$ : trait value after time $t$,

* $x_{opt}$ : optimal (target) trait value,

* $\alpha$ : selection strength (rate of reversion to optimum),

* $\sigma$ : scale parameter of the underlying Gaussian process, controlling the
  magnitude of random fluctuations.

When $\alpha = 0$ , the model reduces to Brownian motion:

$$
x_t \sim \mathcal{N}(x_0,\sigma)
$$

indicating neutral evolution with no selective constraint.

This formulation allows **RPEM** to simulate a wide range of evolutionary
scenarios--from neutral drift to strong stabilizing selection--making it
suitable for generating realistic trait data under complex models, testing
comparative methods, or training machine learning algorithms in evolutionary
biology.

#### Quantitative traits

In **RPEM**, quantitative trait evolution is simulated using the function
`traitEvolSim()`. This function implements a stochastic evolutionary model--by
default, Brownian motion, but extendable to Ornstein-Uhlenbeck and other
processes--and allows users to simulate trait trajectories along evolutionary
branches or edges of a phylogenetic graph.

At a minimum, `traitEvolSim()` requires three arguments:

```{r}
traitEvolSim(
    name = "Neutral",   ## Name of the trait (for labeling and tracking)
    sigma = 2.5,        ## Standard deviation of the process (e.g., mm/Ma)
    step = 0.05         ## Time step for simulation (e.g., in million years)
  ) -> ts_n
```

* `name` : A character string identifying the trait (useful when simulating
  multiple traits).

* `sigma` : The evolutionary rate parameter, representing the standard deviation
  of the trait change per unit time under Brownian motion.

* `step` : The time increment used in the simulation, which should be compatible
  with the time scale of the input phylogenetic graph.

The function returns a simulator object (`ts_n` in this case), which can be used
to evolve trait values across a phylogeny or network. This object contains
methods for recalibrating parameters and simulating trait changes over specified
time intervals.

This modular design supports the simulation of both neutral evolution and
adaptive scenarios (e.g., with stabilizing selection via an optimum), enabling
the generation of realistic trait datasets for method testing, comparative
analysis, or machine learning applications.

When called in this way, $\alpha = 0$, meaning there is no stabilizing
selection, and the trait evolves according to a purely diffusive
process--equivalent to Brownian motion. This model assumes neutral evolution,
where changes accumulate randomly over time, with no directional or stabilizing
forces.

The trait evolution is simulated as follows:

```{r, fig.width = 7.0}
trait <- numeric(1001)      ## For storing the values
trait[1] <- 40              ## Initial trait value

## Update the trait value at time t using the value at time t - 1:
for(i in 2:length(trait))
  trait[i] <- ts_n$updateValue(trait[i - 1])

plot(y = trait, x=0.05*(0:1000), type="l", las=1, ylab="Mean leaf width (mm)",
     xlab="Time (Ma)")
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "trait1",
  paste(
    "Neutral evolution of a quantitative trait (mean leaf width) over a period",
    "of 50 million years."
  )
)

figCounter$getCaption("trait1")
```

This simulation spans 1,001 time steps (50 million years in increments of
$0.05\,\mathrm{Ma}$), starting from an initial value of $40\,\mathrm{mm}$. At
each step, the trait is updated using the `$updateValue()` method of the
simulator object `ts_n`, which draws a new value based on the OU/Brownian motion
transition model. The resulting trajectory illustrates the random walk behavior
characteristic of neutral evolution, where variance in the trait increases with
time.

This approach enables realistic simulation of continuous trait evolution under
user-defined parameters, supporting applications in comparative biology, method
validation, and theoretical modeling.

When the argument $\alpha > 0$, the evolutionary process includes stabilizing
selection toward one or more optimal trait values, and the simulation no longer
follows pure Brownian motion. In this case, the Ornstein-Uhlenbeck (OU) model is
used, and the optima argument must be provided.

For example, when specifying `alpha = 1`, a single optimum can be defined as
follows:

```{r}
traitEvolSim(
  name = "OUsingle",    ## Name of the trait
  alpha = 1,            ## Selection strength (pull toward optimum)
  sigma = 2.5,          ## Evolutionary rate (standard deviation of changes)
  optima = 10,          ## Single optimal trait value
  step = 0.05           ## Time step for simulation (in million years)
) -> ts_OUs
```

* `alpha = 1` indicates a relatively strong attraction toward the optimum.

* `optima = 10` sets the long-term adaptive target for the trait (e.g., an
  optimal leaf width of $10\,\mathrm{mm}$).

The trait will fluctuate around this optimum due to random perturbations scaled
by $\sigma$, but deviations are pulled back proportionally to the distance from the optimum and the value of $\alpha$.

This configuration is useful for simulating traits under stabilizing selection,
such as morphological or physiological characters constrained by natural
selection. The resulting trajectories reflect bounded variation, making them
more biologically realistic for many empirical systems.

Now, the simulator is configured with a single optimum value of
$10\,\mathrm{mm}$. The evolution of the trait is computed in a manner analogous
to the neutral case, but under the influence of stabilizing selection that pulls
the trait toward this optimum.

The simulation proceeds as follows:

```{r, fig.width = 7.0}
trait <- numeric(1001)      ## For storing the values
trait[1] <- 20              ## Initial trait value (far from optimum)

## Update the trait value at time t using the value at time t - 1:
for(i in 2:length(trait))
  trait[i] <- ts_OUs$updateValue(trait[i - 1])

plot(y = trait, x=0.05*(0:1000), type="l", las=1, ylab="Mean leaf width (mm)",
     xlab="Time (Ma)")
abline(h=10, lty=3)
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "trait2",
  paste(
    "Non-neutral evolution of a quantitative trait (mean leaf width) over a",
    "period of 50 million years, with a single trait optimum (10 mm)."
  )
)

figCounter$getCaption("trait2")
```

Starting from an initial value of $20\,\mathrm{mm}$--twice the optimal value--
the trait rapidly evolves toward the optimum ($10\,\mathrm{mm}$) due to the
relatively strong selection rate ($\alpha = 1$). Over time, the trait fluctuates
around the optimum, with deviations dampened by the restoring force of the OU
process. 

This trajectory illustrates the hallmark of stabilizing selection: convergence
toward a phenotypic optimum with bounded variation, in contrast to the unbounded
drift seen under Brownian motion. Such simulations are essential for studying
adaptive evolution, modeling trait optima shifts, and testing comparative
methods that account for selective constraints.

When multiple optima are specified via the optima argument, the trait can shift
between different adaptive peaks over time. To model this behavior, **RPEM**
requires a shift intensity matrix (also known as a transition rate matrix) to be
provided through the transition argument.

This matrix defines the rates at which the evolving system switches from one
optimum to another. Each off-diagonal element $q_{ij}$ represents the
instantaneous rate of shifting from optimum $i$ to optimum $j$. Diagonal
elements are set to `NA` (and internally computed to ensure row sums are zero),
as is standard for rate matrices.

In the example below, three optima are defined ($10$, $15$, and
$20\,\mathrm{mm}$), and shifts between adjacent optima occur at a rate of $0.01$
per million years, while direct shifts between non-adjacent optima (e.g., from
$10$ to $20$) are disallowed (set to $0$):

```{r}
traitEvolSim(
    name = "OUmultiple",  ## Name of the trait
    alpha = 1,            ## Selection strength toward the current optimum
    sigma = 2.5,          ## Evolutionary rate (standard deviation of changes)
    optima = c(10,15,20), ## Three distinct trait optima
    transition = matrix(
      c(NA,   0.01, 0 ,
        0.01, NA,   0.01,
        0,    0.01, NA),
      3, 3),
    step = 0.05           ## Time step for simulation (in million years)
  ) -> ts_OUm
```

* The transition matrix is $3 \times 3$, matching the length of optima.

* A value of 0 indicates no direct transition between two optima.

* The selection strength ($\alpha = 1$) applies to the current optimum: while
  near optimum 10, the trait is pulled toward 10; if a shift occurs to optimum
  15, the pull switches to 15.

This setup enables simulation of adaptive regime shifts, such as those occurring
during colonization of new environments or evolutionary diversification across
niches. It provides a powerful framework for generating realistic trait
evolution scenarios under complex selective landscapes.

The simulation of traits under multiple adaptive optima requires tracking not
only the trait value but also the current evolutionary regime--i.e., which
optimum is currently active. This is managed through a state index, which
represents the current optimum state and evolves according to the transition
rate matrix.

The process involves two key steps at each time point:

1. Update the current optimum using the `$updateState()` method.

2. Simulate the trait value under the current optimum using `$updateValue()`.

This is implemented as follows:

```{r, fig.width = 7.0}
trait <- numeric(1001)      ## For storing the trait values
trait[1] <- 20              ## Initial trait value
state <- integer(1001)      ## For storing the current optimum (state index)
state[1] <- 1               ## Initial optimum (first regime: 10 mm)

## Update the state (optimum) and then the trait value at each time step:
for(i in 2:length(trait)) {
  state[i] <- ts_OUm$updateState(state[i - 1])
  trait[i] <- ts_OUm$updateValue(trait[i - 1], state[i])
}

plot(y = trait, x=0.05*(0:1000), type="l", las=1, ylab="Mean leaf width (mm)",
     xlab="Time (Ma)")
lines(x=0.05*(0:1000), y=c(10,15,20)[state], col="black", lty=3L)
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "trait3",
  paste(
    "Non-neutral evolution of a quantitative trait (mean leaf width) over a",
    "period of 50 million years, with three trait optima (10, 15, and 20 mm)",
    "that are shifting with time."
  )
)

figCounter$getCaption("trait3")
```

* The state vector tracks which optimum (1, 2, or 3) is active at each time
  step, evolving stochastically according to the transition matrix.

* The trait evolves toward the current optimum (e.g., 10, 15, or 20 mm), with
  shifts occurring probabilistically over time.

* The dashed black line (`lty = 3`) shows the trajectory of the instantaneous
  optimum, allowing visual comparison between the trait value and its target.

This approach enables the simulation of complex adaptive landscapes, where
species or populations shift between selective regimes--modeling phenomena such
as niche switching, adaptive radiations, or environmental fluctuations. It
provides a flexible framework for generating realistic evolutionary scenarios
under non-stationary selection.

#### Semi-quantitative or qualitative traits

While the `traitEvolSim()` function in **RPEM** is designed to simulate
continuous, quantitative traits under models such as Brownian motion or
Ornstein-Uhlenbeck, it is straightforward to extend these simulations to
generate semi-quantitative or qualitative traits. This is achieved by
transforming the underlying continuous trait trajectory into discrete states
using biologically informed rules.

Ordered (Semi-Quantitative) Traits
: An ordered categorical trait--such as root system size classified as tiny,
small, normal, large, or huge--can be generated by applying thresholds to a
simulated quantitative trait. These thresholds can be defined based on quantiles
or fixed values.

Binary Traits
: A binary trait (e.g., presence/absence of a feature, smooth vs. dentate leaf
margins) can be derived by applying a single threshold--such as the median of
the simulated trait--where values above the threshold represent one state and
values below represent the other.

Multistate Qualitative Traits
: For unordered multistate traits (e.g., flower color), one approach is to
simulate multiple independent quantitative deviates (one per state), and assign
the trait state at each time point to the deviate with the highest value. This
method models transitions among states as competitive outcomes of underlying
latent evolutionary processes.

The following code demonstrates how to generate an ordered, a binary, and a five-category qualitative trait using simulated neutral evolution trajectories: 

```{r, fig.width = 7.0}
## Generate five random deviates (n = 300 each) using a neutral evolution
## simulator:
traits <- matrix(NA, 300, 5)
traits[1,] <- c(0,0,1,0,0)

for(i in 1:5)
  for(j in 2:nrow(traits))
    traits[j,i] <- ts_n$updateValue(traits[j - 1,i])

## Simulate an ordered trait:
bnd <- quantile(traits[,1],seq(0,1,length.out=6))
bnd[c(1L,6L)] <- c(-Inf,Inf)
cut(
  traits[,1],
  bnd,
  c("tiny","small","normal","large","huge"),
  ordered=TRUE
) -> var_ordered

pp <- par(no.readonly = TRUE)
par(mar=c(5,7,2,2))
plot(y=var_ordered, x=seq(0,50,length.out=length(var_ordered)), type="l",
     yaxt="n", xlab="time (Ma)", ylab="Root system size\n")
axis(2, at=1:5, labels=levels(var_ordered), las=1)
par(pp)
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "trait4",
  paste(
    "Evolution of a semi-quantitative trait (size of the plant's root system",
    "represented as five categories) over a period of 50 million years. The",
    "sifts are modeled using a quantitative trait evolving neutrally."
  )
)

figCounter$getCaption("trait4")
```

```{r, fig.width = 7.0}
## Simulate a binary trait:
traits[,1] %>%
  {. > median(.)} -> var_binary
pp <- par(no.readonly = TRUE)
par(mar=c(5,7,2,2))
plot(y=var_binary, x=seq(0,50,length.out=length(var_binary)), type="l",
     yaxt="n", xlab="time (Ma)", ylab="Leaf margin\n")
axis(2, at=c(0,1), labels=c("Smooth","Dentate"), las=1)
par(pp)
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "trait5",
  paste(
    "Evolution of a binary trait (leaf margin either smooth or dentate) over a",
    "period of 50 million years. The sifts are modeled using a quantitative",
    "trait evolving neutrally."
  )
)

figCounter$getCaption("trait5")
```

```{r, fig.width = 7.0}
## Simulate a qualitative trait:
traits %>%
  apply(
    1L,
    which.max
  ) %>%
  c("blue","green","red","yellow","purple")[.] %>%
  as.factor -> var_qualitative

pp <- par(no.readonly = TRUE)
par(mar=c(5,7,2,2))
plot(y=var_qualitative, x=seq(0,50,length.out=length(var_qualitative)),
     type="l", yaxt="n", xlab="time (Ma)", ylab="Flower color\n")
axis(2, at=1:5, labels=levels(var_qualitative), las=1)
par(pp)
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "trait6",
  paste(
    "Evolution of a quantitative trait (the color of flowers represented as",
    "five colors) over a period of 50 million years. The sifts are modeled",
    "using a quantitative trait evolving neutrally. It thus entails that each",
    "of them is to proceed in a step wise manner with respect to the order",
    "going from blue to green, to red, to yellow, and then to purple, in",
    "either direction."
  )
)

figCounter$getCaption("trait6")
```

This approach allows users to simulate realistic discrete traits that reflect
complex evolutionary dynamics, while leveraging the flexibility of
continuous-time stochastic models. These simulated traits can be used to test
phylogenetic comparative methods, model character evolution, or generate
training data for machine learning applications.

[RENDU ICI...]

### volutionary Network Simulation 

While the simulator generator functions `molEvolSim()` and `traitEvolSim()`
enable the simulation of DNA sequence and quantitative trait evolution, these
processes must ultimately be applied across evolutionary graphs of varying
complexity--ranging from simple phylogenetic trees to reticulated phylogenetic
networks. To support such analyses, RPEM provides the `randomGraph()` function,
a flexible tool for generating random evolutionary graphs under customizable
rules.

The `randomGraph()` function builds graphs incrementally using a vertex addition
process, starting from a single origin (root) vertex and adding new vertices one
at a time. Each new vertex is connected to existing vertices in the graph
according to user-defined probabilistic rules that govern:

* The number of outgoing edges (children) per vertex,

* The number of incoming edges (parents) for each new vertex,

* The length of edges (evolutionary distance),

* Constraints on reticulation (e.g., limiting lateral transfer between distant
  lineages).

**Key Arguments of `randomGraph()`**

* `NV` : The total number of vertices to generate.

* `NC` : A function that returns the maximum number of children (outgoing edges)
  for a vertex when it is created. This can be constant or randomly drawn.

* `NP` : A function that returns the number of parents (incoming edges) for a
  new vertex. Must be at least 1. Setting `NP = function(...) 1` yields a tree;
  higher values allow reticulation.

* `timestep` : A function that returns the edge length (evolutionary distance)
  for newly created edges, often sampled from a distribution (e.g., uniform or
  exponential).

* `maxDist` : A function that returns the maximum allowed distance between
  potential parent vertices. If two candidate parents are separated by a
  distance greater than this value, they cannot both be parents of the same new
  vertex--this limits biologically implausible lateral transfers.
  
* `...` : Additional named arguments passed to the functions `NC`, `NP`,
  `timestep`, and `maxDist`. It is recommended that these functions include a
  `...` argument in their definition to accept (and ignore) any extra
  parameters.

When a vertex reaches zero remaining child capacity (due to `NC` being
decremented with each new outgoing edge), no further edges can originate from
it.

#### Example: Simulating a Dichotomous Tree 

The following code generates a binary (dichotomous) phylogenetic tree with 50
vertices:

```{r, fig.height = 5, fig.width=7.15}
randomGraph(
  NV = 50,
  NC = function(...) 2,  ## Each vertex has exactly two children
  NP = function(...) 1,  ## Each new vertex has one parent (tree structure)
  timestep = function(ts_min, ts_max, ...) runif(1, ts_min, ts_max),
  maxDist = function(...) NULL,
  ts_min = 2,
  ts_max = 5,
  verbose = FALSE
) -> gr_dich

gr_dich

plot(gr_dich, cex.min=3, cex.lab=0.5)
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "randomGraph1",
  paste(
    "A 50-vertex random dichotmic tree generated using function randomGraph()."
  )
)

figCounter$getCaption("randomGraph1")
```

This example produces a strictly bifurcating tree, where every internal vertex
has exactly two descendants, and all edges have lengths drawn uniformly between
2 and 5 units of evolutionary time.

This modular and extensible design allows `randomGraph()` to simulate a wide
range of evolutionary scenarios--from pure vertical descent (trees) to complex
reticulate evolution (networks)--making it a powerful tool for generating
synthetic phylogenetic structures for method development, simulation studies,
and machine learning applications in evolutionary biology.

The functions `NC` and `NP` are passed arbitrary arguments such as `ts_min` and
`ts_max` by `randomGraph()`, even though they do not use them. This is because
`randomGraph()` uniformly passes all additional named arguments (via `...`) to
each of the user-defined functions--`NC`, `NP`, `timestep`, and `maxDist`--for
simplicity and consistency in interface design.

In this example:

* `NC = function(...) 2` and `NP = function(...) 1` ignore `ts_min` and `ts_max`
because they return fixed values, but they are still compatible because they
accept `...`.

* `timestep = function(ts_min, ts_max, ...) runif(1, ts_min, ts_max)` explicitly
  uses `ts_min` and `ts_max` to generate edge lengths from a uniform
  distribution.

* `maxDist = function(...) NULL` consistently returns `NULL`, effectively
  disabling the constraint on parent dissimilarity.

Indeed, since `NP returns 1` in this case (i.e., each new vertex has only one
parent), the graph remains a tree--no reticulation occurs. As a result, the
`maxDist` function is never meaningfully invoked, because the condition it
governs (distance between multiple parents) does not arise. When a vertex has
only one parent, there is no need to evaluate the distance between parents, so
`maxDist()` is bypassed in practice.

This illustrates an important point: `maxDist` only becomes relevant when `NP`
returns values greater than 1, enabling reticulate evolution (e.g.,
hybridization or lateral gene transfer). In such cases, multiple candidate
parents may exist, and `maxDist` ensures that they are not too distantly
separated in the graph--enforcing biological realism in lateral transfer events.

Thus, while `ts_min`, `ts_max`, and `maxDist` are defined in the call, they are
either unused or inactive under the current configuration. This highlights the
importance of aligning the behavior of these functions with the desired graph
structure—tree-like vs. reticulated.

Another way to generate a random phylogenetic tree is to use the `rtree()`
function from the **ape** package, a widely used tool for simulating and
manipulating phylogenetic trees under a pure birth (Yule) or random branching
process.

For instance:

```{r, fig.height = 5, fig.width=7.15}
rtree(25) %>%
  as.graph -> gr_rtree

plot(gr_rtree, cex.min=1, show.vertex.labels = FALSE)
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "randomGraph2",
  paste(
    "A 25-tip random dichotmic tree generated using function rtree() from",
    "package ape that has been converted into a (49-vertex) phylogenetic graph."
  )
)

figCounter$getCaption("randomGraph2")
```

Here: 

* `rtree(25)` generates a random bifurcating tree with 25 tips (extant species
  or operational taxonomic units).
  
* The resulting "phylo" object is converted into a phylogenetic graph using
  `as.graph()`, a function from **RPEM** that transforms tree structures into a
  graph representation compatible with downstream simulation steps.
  
* The plot displays the resulting tree with vertex size scaled by node type
  (e.g., tips vs. internal nodes), without labeling individual vertices.

While this approach provides a quick and biologically plausible tree topology
for use in simulation studies, especially when the focus is on tree-like
evolution rather than reticulate processes, `randomGraph()` offers fine-grained
control over graph growth dynamics.

Finally, a random reticulated phylogenetic graph is generated as follows:

```{r, fig.height = 5, fig.width=7.15}
randomGraph(
  NV = 50,
  NC = function(lambda_child, ...) 1 + rpois(1, lambda_child),
  NP = function(lambda_parent, ...) 1 + rpois(1, lambda_parent),
  timestep = function(ts_min, ts_max, ...) runif(1, ts_min, ts_max),
  maxDist = function(max_anc, ...) runif(1, 0, max_anc),
  lambda_child = 3.0,
  lambda_parent = 3.0,
  ts_min = 5,
  ts_max = 12,
  max_anc = 13,
  verbose = FALSE
) -> gr_ret

gr_ret

plot(gr_ret, cex.min=3, cex.lab=0.5)
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "randomGraph3",
  paste(
    "A 50-vertex reticulated phylogenetic graph generated using randomGraph()",
    "in the **RPEM** package. This network represents a non-tree-like",
    "evolutionary history, where reticulation edges (e.g., hybridization or",
    "lateral gene transfer events) connect vertices that have more than one",
    "parent. The graph was constructed by incrementally adding vertices and",
    "allowing multiple incoming edges per vertex, with edge lengths",
    "representing evolutionary time. For visualization, the network is",
    "displayed as a tree-like structure (black branches) with reticulation",
    "events overlaid as red arrows, highlighting the locations of lateral",
    "evolutionary contributions."
  )
)

figCounter$getCaption("randomGraph3")
```

In this simulation: 

* The number of children per vertex (outgoing edges) is set to
  $NC \sim 1 + \mathcal{Poisson}(\lambda=lambda\_child)$, ensuring each vertex
  has at least one child and an average of 4.0.

* The number of parents per new vertex (incoming edges) is
  $NP \sim 1 + \mathcal{Poisson}(\lambda=lambda\_parent)$, allowing for
  reticulation (e.g., hybridization or lateral gene transfer), with an average
  of 4.0 parents per vertex.

* Edge lengths (evolutionary distances) are drawn uniformly from the interval
  $[ts\_min,ts\_max] = [5,12]$.

* The `maxDist` function constrains reticulation events by requiring that the
  distance between potential parent vertices be no greater than a random value
  drawn uniformly from the interval $[0,max\_anc] = [0,13]$. This enforces
  biological plausibility by limiting lateral transfers between highly divergent
  lineages.

The resulting graph contains both vertical descent and reticulate connections,
representing a phylogenetic network rather than a strictly bifurcating tree.

For visualization, the reticulated graph is first coerced into a tree-like
structure using the `as.phylo` method, which retains a primary treelike
scaffold. The edges that are discarded during this coercion--i.e., the
additional reticulation edges that violate the tree structure--are then overlaid
as red arrows on top of the black tree segments. This highlights the locations
and directions of lateral evolutionary events, providing a clear visual
representation of network complexity.

This approach enables the simulation and visualization of realistic evolutionary
histories involving hybridization, horizontal gene transfer, or
recombination--making **RPEM** a powerful tool for generating synthetic data to
test network inference methods, comparative models, and machine learning
algorithms in evolutionary genomics.

### Simulating DNA Sequences and Trait Evolution Over a Graph

We can now simulate DNA sequence evolution and trait dynamics directly on the
random evolutionary graphs generated earlier, using the **RPEM** functions
`simulateSequence()` and `simulateTrait()`. These functions propagate molecular
or phenotypic change across the graph by treating edge lengths as evolutionary
time intervals.

The simulation proceeds from ancestral to descendant vertices, applying
stochastic evolutionary models (e.g., nucleotide substitution or
Ornstein-Uhlenbeck trait change) along each edge. In reticulated graphs, special
handling is required for vertices with multiple parents (i.e., hybridization or
lateral transfer events), which is managed through user-defined parental
contribution weighting.

#### Simulating DNA Sequence Evolution

Function `simulateSequence()` simulates the evolution of a DNA sequence along
the structure of a phylogenetic graph, including reticulated networks. It is
called as follows:

```{r, fig.width=7.15}
gr_ret %>%
  simulateSequence(
    Q = DNArate(model="K80", piGap=0.25, insertionRate=0.15, deletionRate=0.15),
    sqn = drawDNASequence(NN=200, piGap=0.25, pi=c(0.3,0.2,0.2,0.3)),
    rate = drawEvolRate(NN=200, gamma.shape=4, gamma.scale=4e-4),
    a = 1
  ) -> seq_ret

seq_ret %>%
  concatenate(discard="-") %>%
  show.sequence
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "graphSeq",
  paste(
    "A set of 50 DNA sequences generated by evolving a randomly generated",
    "ancestral sequence along the 50-vertex reticulated phylogenetic graph",
    "(Figure 11) using the simulation framework in RPEM. Each sequence",
    "corresponds to a terminal vertex (tip) in the network, and evolution was",
    "modeled under a K80 substitution model with gamma-distributed rate",
    "variation across sites and a uniform gap frequency of 25%. The color",
    "coding is as follows: red = Adenine (A), yellow = Guanine (G),",
    "green = Cytosine (C), and blue = Thymine (T). This alignment represents a",
    "synthetic dataset that includes both vertical descent and reticulate",
    "evolutionary events, making it suitable for benchmarking phylogenetic",
    "network inference methods or training machine learning models in",
    "evolutionary genomics."
  )
)

figCounter$getCaption("graphSeq")
```

Function arguments are:

* `x` : A graph-class object (e.g., `gr_ret`) representing the phylogenetic or
  reticulated network.

* `Q` : A shift rate matrix generated by `DNArate()`, specifying substitution
  and indel rates.
  
* `sqn` : An initial DNA sequence, typically generated via `drawDNASequence()`,
  stored as a raw vector.

* `rate` : A numeric vector of site-specific evolutionary rates, usually
  generated by `drawEvolRate()`, allowing for among-site rate variation.

* `a` : A parameter controlling the weighting function used to combine genetic
  contributions from multiple parent vertices in reticulate nodes.

* `weighting` : (Optional) A custom function that computes the relative
  contribution of each parent based on edge distances. If not provided, the
  default inverse-distance weighting is used.

* `...` : Additional arguments passed internally to the weighting function.

#### Parental Contribution Weighting in Reticulate Nodes 

In reticulated graphs, a vertex may have more than one parent due to
hybridization, horizontal gene transfer, or recombination. To compute the state
of such a vertex, contributions from all parents must be combined using a
weighting function $w_a(d_i)$, which assigns weights based on the distances
$d_i$ from each parent to the child.

The default weighting function is: 

$$
w_a(d_i) = \frac{d_i^{-a}}{\sum_{i=1}^n d_i^{-a}},
$$

where:

* $d_i$ is the length of the edge from parent $i$,

* $a$ is a user-defined exponent controlling how strongly closer parents are
  favored.

Special cases:

* $a = 0$ : all parents contribute equally (distance-independent),

* $a = 1$ : weights are proportional to the inverse of the edge distance, and

* $a = 2$ : weights are proportional to the inverse squared distance (stronger
  bias toward closer parents).

In the example above, $a = 1$ implies that parental contributions are inversely
proportional to their edge lengths--shorter edges (more direct evolutionary
influence) receive higher weight.

Users may supply a custom weighting function to implement alternative schemes
(e.g., equal splitting, time-thresholded contributions, or biological models of
recombination).

This framework enables biologically realistic simulation of non-tree-like
evolution, supporting applications in phylogenetic network inference, horizontal
gene transfer detection, and machine learning model training.

#### Simulating Trait Evolution 

The `simulateTrait()` function in RPEM allows for the simulation of quantitative
trait evolution across both tree-like and reticulated phylogenetic graphs. It
applies a specified evolutionary model--such as Brownian motion or
Ornstein-Uhlenbeck--along the edges of the graph, using edge lengths as
evolutionary time intervals.

For vertices with multiple parents (i.e., reticulate nodes), the trait value is
computed as a weighted average of the contributions from all parent vertices,
where weights are determined by the a parameter and the edge distances (as in
`simulateSequence()`).

**Neutral Trait Evolution (Brownian Motion)**

To simulate neutral evolution, we use a Brownian motion model implemented via
`traitEvolSim()`. The following code simulates a quantitative trait (e.g., mean
leaf width) across the reticulated graph `gr_ret`:

```{r here, fig.height=5, fig.width=7}
gr_ret %>%
  simulateTrait(
    tem = traitEvolSim("Neutral", sigma=1),
    value = 20,
    a = 1
  ) -> trait_ret_n

trait_ret_n$value

col <- head(rainbow(1200), 1000)

plot(gr_ret, y=trait_ret_n$value, bg=col, cex.min=3, cex.lab=0.5, ylim=c(5,35))

legend(x=40, y=32, pch=21, legend=seq(5,35,6), title="Width (mm)",
       pt.bg=col[1 + ceiling((seq(5,35,6) - 5)/(35-5)*999)], xpd=TRUE)
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "graphTrait1",
  paste(
    "Neutral evolution of a quantitative trait (mean leaf width) shown at the",
    "vertices of a phylogenetic graph. The trait was simulated under a",
    "Brownian motion model across a 50-vertex reticulated phylogenetic network",
    "generated by `randomGraph()`. Edge lengths represent evolutionary time,",
    "and trait values were propagated from ancestor to descendant vertices,",
    "with changes accumulating proportionally to the square root of time and",
    "the evolutionary rate parameter $\\sigma$. At reticulation events,",
    "parental contributions were combined using inverse-distance weighting",
    "($a = 1$). The resulting variation in leaf width reflects unconstrained",
    "divergence, characteristic of neutral evolution, with values ranging",
    "widely across the graph."
  )
)

figCounter$getCaption("graphTrait1")
```

Here: 

* `tem` specifies the trait evolution model: neutral (Brownian motion) with
  evolutionary rate sigma = 1,

* `value = 20` sets the initial trait value at the root, and

* $a = 1$ uses inverse-distance weighting to combine trait values from multiple
  parents at reticulation events.

The resulting trait values show substantial variation across the graph,
reflecting the unconstrained, random-walk nature of Brownian motion.

**Non-Neutral Trait Evolution (Ornstein-Uhlenbeck with Single Optimum)**

To simulate stabilizing selection, we apply an Ornstein-Uhlenbeck model with a
single optimum: 

```{r there, fig.height=5, fig.width=7}
gr_ret %>%
  simulateTrait(
    tem = traitEvolSim("OUsingle", 1, alpha=1, optima=c(10)),
    value = 10,
    a = 1
  ) -> trait_ret_OUs

plot(gr_ret, y=trait_ret_OUs$value, bg=col, cex.min=3, cex.lab=0.5,
     ylim=c(7,13))

legend(x=40, y=32, pch=21, legend=seq(7,13,1), title="Width (mm)",
       pt.bg=col[1 + ceiling((seq(7,13,1) - 7)/(13-7)*999)], xpd=TRUE)
```
```{r, echo=FALSE, results='asis'}
figCounter$register(
  "graphTrait2",
  paste(
    "Non-neutral evolution (single optimum of 10 mm) of a quantitative trait",
    "(mean leaf width) shown at the vertices of a phylogenetic graph. The",
    "trait was simulated under an Ornstein-Uhlenbeck (OU) model across a",
    "50-vertex reticulated phylogenetic network, where evolutionary",
    "trajectories are drawn toward a stabilizing optimum",
    "($x_{opt} = 10\\,\\mathrm{mm}$) with selection strength $\\alpha = 1$.",
    "Edge lengths represent evolutionary time, and trait values at",
    "reticulation events are computed using inverse-distance weighting",
    "(`a = 1`). Compared to the neutral model (Figure 13), the trait exhibits",
    "much smaller variation, as deviations from the optimum are corrected over",
    "time--reflecting the action of stabilizing selection."
  )
)

figCounter$getCaption("graphTrait2")
```

In this case: 

* $alpha = 1$ indicates strong attraction toward the optimum (optima = 10 mm),

* The trait starts at the optimum and remains tightly clustered around it due to
  stabilizing selection.

#### Comparison of Evolutionary Models

The simulated trait exhibits much smaller variation under the non-neutral (OU)
model compared to the neutral (Brownian motion) model. This reduced variance
arises because the OU process continuously pulls the trait toward its optimal
value, mimicking the effects of stabilizing selection. In contrast, the Brownian
motion model allows trait values to drift freely, resulting in greater
divergence over time. 

These simulations demonstrate how **RPEM** can be used to explore the effects of
different evolutionary regimes--neutral drift versus adaptive constraints--on
trait distribution across complex phylogenetic graphs, including those with
reticulation. This capability is essential for testing comparative methods,
modeling adaptive landscapes, and generating training data for machine learning
in evolutionary biology.

## Conclusion 

The RPEM package provides--in addition to phylogenetic eigenvector maps
calculations--a comprehensive and flexible framework for simulating evolutionary
processes on both tree-like and reticulated phylogenetic graphs. Through
functions like `randomGraph()`, `molEvolSim()`, and `traitEvolSim()`, RPEM
enables users to:

* Generate random phylogenetic trees and networks under customizable growth
  rules.

* Simulate DNA sequence evolution with substitution, indel, and among-site rate
  variation.

* Model quantitative trait evolution under neutral (Brownian motion) and
  adaptive (Ornstein-Uhlenbeck) regimes.

* Simulate semi-quantitative and qualitative traits by transforming underlying
  continuous traits,

* Propagate evolutionary change across complex graphs, including reticulate
  events such as hybridization or lateral gene transfer.

These capabilities make **RPEM** a powerful tool for:

* testing and benchmarking phylogenetic inference methods,

* developing and training machine learning algorithms for phylogenomic analysis,

* exploring theoretical evolutionary dynamics under controlled conditions, and

* generating realistic synthetic datasets when empirical data are limited.

By integrating graph theory, stochastic modeling, and phylogenetics, **RPEM**
bridges the gap between theoretical models and empirical applications, offering
researchers a robust platform for method development and hypothesis testing in
evolutionary biology.

We invite users to explore the full potential of RPEM, contribute to its 
development, and apply it to novel questions in evolutionary genomics and
beyond.

## References
